---
title: "Supplementary Material"
date: "2026-01-28"
editor: visual
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 4
execute:
  eval: true
  echo: false
  cache: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache.lazy = FALSE)
```



## Supplemenentary methods
#### Bayesian spatial modelling of bleaching severity

We modelled aerial bleaching severity as an ordinal response using a Bayesian
spatial proportional-odds model fitted with the integrated nested Laplace
approximation (INLA; Rue et al. 2009) and the stochastic partial differential
equation (SPDE) approach to spatial random fields (Lindgren et al. 2011), as
implemented in the R-INLA package (www.r-inla.org).

#### Response variable

Reef-level aerial bleaching scores were collapsed from the six-category survey
scale into four ordered classes to harmonise observations across survey eras
(1998–2002, 2016–2020, 2022–2025) and to ensure each class was adequately
represented: class 1 (no bleaching; original score 0), class 2 (mild; scores
1–2), class 3 (moderate; score 3) and class 4 (severe; scores 4–5). The ordinal
structure was modelled with a proportional-odds (cumulative logit) likelihood
(`family = "pom"`), which estimates a common set of covariate effects and
class-specific cut-points (thresholds) on the latent logit scale. We verified
that the estimated cut-points were strictly ordered after back-transformation
from INLA's internal increment parameterisation (α1 = θ1; αk = α(k−1) + exp(θk)),
confirming a valid cumulative ordering (α = −2.57, 0.46, 1.82).

#### Linear predictor

The latent linear predictor for reef i comprised a fixed-effect component, a
continuous spatial random field, and a reef-level independent random effect:

`η_i = β0 + β_DHW · DHW_i + β_year[t(i)] + β_int[t(i)] · DHW_i
+ u(s_i) + v_i`

where DHW is the annual maximum Degree Heating Weeks experienced by the reef in
the survey year, year is a categorical fixed effect for the eight mass-bleaching
events (1998, 2002, 2016, 2017, 2020, 2022, 2024, 2025; 1998 as reference), and
the DHW × year interaction allows the dose–response slope to differ among events.
The year main effects (β_year) estimate the contextual bleaching propensity at
reference thermal stress, and the interaction terms (β_int) estimate event-
specific modulation of the thermal dose–response. u(s_i) is the spatial field
(below) and v_i a zero-mean Gaussian reef-level random effect (iid) capturing
residual reef-specific heterogeneity.

##### Spatial random field (SPDE)

The spatial field u(s) was modelled as a zero-mean Gaussian random field with a
Matérn covariance, approximated by the SPDE–GMRF representation on a constrained
triangulated mesh. The study domain was defined by a concave hull
(concaveman, concavity = 2) of the Great Barrier Reef Marine Park reef
features, buffered by 10 km and simplified (10 km tolerance) to a smooth outer
boundary. All spatial operations used the projected coordinate system UTM Zone
55S (EPSG:32755), so all distances and ranges are reported in metres.

The mesh was constructed with inla.mesh.2d using the domain boundary, an inner
maximum edge length of 40 km and an outer (extension) edge of 800 km, with an
offset of 5 km / 10 km for the inner and outer boundaries respectively. The
Matérn smoothness was fixed at ν = 1 (α = 2). Observation and prediction
locations were projected onto the mesh with the corresponding sparse projector
matrices, and the spatial index was constructed with inla.spde.make.index.
Observation and prediction data were combined in a single estimation stack
(inla.stack).

We placed penalised-complexity (PC) priors on the Matérn range and marginal
standard deviation (Fuglstad et al. 2019; inla.spde2.pcmatern):
P(range < 150 km) = 0.5 and P(σ > 3) = 0.01. PC priors shrink the field toward
the simpler base model (no spatial effect / infinite range) unless the data
support additional complexity, providing conservative, interpretable control of
the spatial term.

##### Priors on fixed effects

We used weakly informative Gaussian priors on all fixed effects
(mean 0, precision 0.001) except the intercept, for which we used a mildly
informative prior (mean −2, precision 1) encoding the prior expectation of low
baseline severe-bleaching probability at zero thermal stress. The cut-point
hyperparameters used INLA's default priors, which yielded well-identified,
strictly ordered thresholds.

##### Inference

Posterior inference used INLA's default integration strategy with a variational-
Bayes correction for the latent field (`control.vb`). We retained model
configurations (`config = TRUE`) to enable posterior sampling of the fixed
effects via `inla.posterior.sample`, drawing from the joint posterior to
propagate the full covariance among parameters when deriving quantities that
combine multiple coefficients (e.g. event-specific dose–response curves and
their thresholds). Model fit was assessed with the Deviance Information Criterion
and the Watanabe–Akaike Information Criterion.

##### Event-specific response curves
To visualise how the bleaching response to thermal stress differed among events, we computed the full posterior distribution over ordinal classes across the observed DHW gradient (0–16 °C-weeks) for each event year. For a given linear predictor η, the probability of each class k was obtained from the cumulative logit form, `P(Y = k) = F(α_k − η) − F(α_{k−1} − η)`, with F the logistic CDF and α the ordered cut-points reconstructed from the model's increment-parameterised thresholds `(α_1 = θ_1; α_k = α_{k−1} + exp(θ_k))`. Linear predictors and their 95% credible bounds were taken from the model's posterior on the latent scale and propagated through this transformation. Curves were truncated at each event's maximum observed DHW so that predictions were not extended beyond the sampled thermal range; the 2025 event, surveyed only above ~7.5 °C-weeks, was treated as having no low-DHW support and is shown accordingly.

##### Spatial prediction
Spatially continuous predictions of bleaching probability were generated on a hexagonal grid (cell size 27.8 km) covering the reef domain. For each grid cell and event year we extracted the area-weighted maximum DHW and projected the cell centroids onto the fitted mesh, then predicted the latent linear predictor at these locations by appending the prediction grid to the estimation stack and re-running INLA with the hyperparameters fixed at the fitted posterior mode (control.mode), so that prediction did not re-estimate the model. The probability of severe bleaching was mapped as `P(Y = severe) = 1 − F(α_3 − η)` per cell, with the posterior mean displayed as the predicted surface. Because this probability is a saturating transformation of the latent field, we report predictive uncertainty on the latent (linear-predictor) scale — the 95% credible interval width of η — which reflects spatial information (mesh precision and survey density) without the additional curvature that the logistic transformation imposes near probabilities of 0.5. Grid-cell predictions carry substantial uncertainty; we therefore interpret the predicted surfaces at the regional scale, consistent with the spatial field's range, rather than at the level of individual cells.

##### Sensitivity analyses

Because the spatial range was a non-trivial fraction of the domain extent, and
because the fixed-effect priors could in principle influence the year-level
estimates that underpin our main inference, we conducted two pre-registered
sensitivity analyses.

_Spatial prior_. We compared the PC-prior specification above against a
diffuse Matérn prior (`inla.spde2.matern` with default internal priors). The
estimated practical range was robust to this choice (PC-prior posterior median
≈ 357 km, 95% credible interval ≈ 219–530 km; diffuse-prior median ≈ 440 km),
indicating that the regional-scale spatial structure is data-driven rather than
an artefact of an unconstrained prior. The fixed-effect estimates were
effectively unchanged between specifications (differences in the fourth decimal
place), confirming that the inferred year-level effects are insensitive to the
spatial prior.

_Fixed-effect prior on the thermal covariate_. We compared a mildly
informative prior on the DHW coefficient (mean 1, precision 1) against a weakly
informative prior (mean 0, precision 0.001). The qualitative result — a
monotonic increase in the year-level intercepts across recent events, with all
recent-event credible intervals excluding zero — was invariant to this choice.
The absolute magnitudes were sensitive to the DHW prior (e.g. the DHW main
effect shifted from 0.61 to 0.82, and recent-year intercepts by ≈ 1.1–1.4 logit
units), reflecting correlation between the thermal main effect and the year
intercepts. We therefore adopt the weakly informative prior as the primary
specification, as it is the more standard and defensible choice and yields
directly interpretable estimates, and we report both fits.

#### Reported spatial parameters

Under the primary model, the spatial field operated at a regional scale
(posterior median practical range ≈ 357 km, 95% CrI ≈ 219–530 km) with a
substantial marginal standard deviation (≈ 3.1 logit units, 95% CrI ≈ 2.2–4.3),
indicating a strong but spatially broad latitudinal gradient in bleaching
propensity rather than fine reef-scale autocorrelation. The reef-level
independent random effect had negligible variance once the spatial field and
fixed effects were included (precision posterior median ≈ 1.3 × 10^4;
variance ≈ 8 × 10^−5), indicating little residual reef-specific heterogeneity
beyond the regional gradient, thermal dose and event effects. We report year
effects net of this validated spatial structure.

Analyses were conducted in `R` using `R-INLA` (vXX), `fmesher`, `sf`,
`terra` and `exactextractr` for spatial data handling and DHW extraction, and
`concaveman` for boundary construction.


```{r}
#| message: false
#| warning: false
#| cache: true

library(tidyverse)
library(terra)
library(sf)
library(dhw)
library(patchwork)

### load all datasets:
GBR_CRW_SST <- rast("~/GBR-dhw/datasets/GBR_CoralTemp_SST.tif")
GBR_CRW_DHW <- rast("~/GBR-dhw/datasets/GBR_CoralTemp_DHW.tif")

GBR_OISST2_SST <- rast("~/GBR-dhw/datasets/GBR_OISST2_SST.tif")
# GBR_OISST2_SSTA <- rast("~/GBR-dhw/datasets/GBR_OISST2_SSTA.tif")
# GBR_OISST2_HS <- rast("~/GBR-dhw/datasets/GBR_OISST2_HS.tif")
GBR_OISST2_DHW <- rast("~/GBR-dhw/datasets/GBR_OISST2_DHW.tif")

GBR_OISST_SST <- rast("~/GBR-dhw/datasets/GBR_OISST_SST.tif")
GBR_OISST_SSTA <- rast("~/GBR-dhw/datasets/GBR_OISST_SSTA.tif")
# GBR_OISST_HS <- rast("~/GBR-dhw/datasets/GBR_OISST_HS.tif")
GBR_OISST_DHW <- rast("~/GBR-dhw/datasets/GBR_OISST_DHW.tif")

GBR_OISST2_SST <- rast("~/GBR-dhw/datasets/GBR_OISST2_SST.tif")
GBR_OISST2_SSTA <- rast("~/GBR-dhw/datasets/GBR_OISST2_SSTA.tif")
# GBR_OISST2_HS <- rast("~/GBR-dhw/datasets/GBR_OISST2_HS.tif")
GBR_OISST2_DHW <- rast("~/GBR-dhw/datasets/GBR_OISST2_DHW.tif")


# gbr_reef_outline <- download_gbr_spatial(return="base")
# gbr_reefs <- download_gbr_spatial(return="combined")
# saveRDS(gbr_reef_outline, "~/GBR-dhw/datasets/gbr_reef_outline.rds")
# saveRDS(gbr_reefs, "~/GBR-dhw/datasets/gbr_reefs.rds")

gbr_reef_outline <- readRDS("~/GBR-dhw/datasets/gbr_reef_outline.rds")
gbr_reefs <- readRDS("~/GBR-dhw/datasets/gbr_reefs.rds")

aus_outline <- gbr_reef_outline |> slice_max(n=1, order_by=SHAPE_AREA)
ausmap <- rnaturalearth::ne_countries(scale="large", country="Australia") |> 
  st_transform(4326) |> 
  st_crop(GBR_OISST_SSTA |> st_bbox()) 


### calculate breakpoints:

library(terra)
library(Rbeast)
library(tidyverse)

GBR_OISST2_SST <- rast("~/GBR-dhw/datasets/GBR_OISST2_SST.tif")

sst_ts <- terra::global(GBR_OISST2_SST, mean, na.rm = TRUE) %>%
  mutate(date = as.Date(rownames(.))) |>
  mutate(
    year  = year(date),
    month = month(date),
    austral_year = if_else(month == 12L, year + 1L, year)  # Dec belongs to next year
  ) %>%
  filter(month %in% seq(1:3)) %>%
  filter(austral_year > 1940, austral_year < 2026) %>%
  group_by(austral_year) %>%
  summarise(sst = mean(mean, na.rm = TRUE), .groups = "drop") %>%
  rename(year = austral_year)

sst_ts <- sst_ts %>% arrange(year)
y_sst <- sst_ts$sst
t_sst <- sst_ts$year

beast_sst <- beast(
  y_sst,
  start  = t_sst[1],
  deltat = 1,
  season = "none",
  print.progress = FALSE,
  quiet = TRUE
)

cp_sst   <- beast_sst$trend$cp
cppr_sst <- beast_sst$trend$cpPr
if (is.null(cppr_sst)) cppr_sst <- beast_sst$trend$cpProb

sst_cp_tbl <- tibble(cp = cp_sst, cp_prob = cppr_sst) %>%
  mutate(year = as.integer(round(cp))) %>%
  arrange(desc(cp_prob))

breakpoints_sst <- sst_cp_tbl$cp[1:min(2, nrow(sst_cp_tbl))]
breakyears_sst  <- sst_cp_tbl$year[1:min(2, nrow(sst_cp_tbl))]

GBR_OISST2_SSTA <- rast("~/GBR-dhw/datasets/GBR_OISST2_SSTA.tif")

ssta_ts <- terra::global(GBR_OISST2_SSTA, mean, na.rm = TRUE) %>%
  mutate(date = as.Date(rownames(.))) |>
  mutate(
    year  = year(date),
    month = month(date),
    austral_year = if_else(month == 12L, year + 1L, year)  # Dec belongs to next year
  ) %>%
  filter(month %in% c(12, 1, 2, 3)) %>%
  filter(austral_year > 1940, austral_year < 2026) %>%
  group_by(austral_year) %>%
  summarise(ssta = mean(mean, na.rm = TRUE), .groups = "drop") %>%
  rename(year = austral_year)


ssta_ts <- ssta_ts %>% arrange(year)
y_ssta <- ssta_ts$ssta
t_ssta <- ssta_ts$year

beast_ssta <- beast(
  y_ssta,
  start  = t_ssta[1],
  deltat = 1,
  season = "none",
  print.progress = FALSE,
  quiet = TRUE
)

cp_ssta   <- beast_ssta$trend$cp
cppr_ssta <- beast_ssta$trend$cpPr
if (is.null(cppr_ssta)) cppr_ssta <- beast_ssta$trend$cpProb

ssta_cp_tbl <- tibble(cp = cp_ssta, cp_prob = cppr_ssta) %>%
  mutate(year = as.integer(round(cp))) %>%
  arrange(desc(cp_prob))

breakpoints_ssta <- ssta_cp_tbl$cp[1:min(2, nrow(ssta_cp_tbl))]
breakyears_ssta  <- ssta_cp_tbl$year[1:min(2, nrow(ssta_cp_tbl))]


GBR_OISST2_DHW <- rast("~/GBR-dhw/datasets/GBR_OISST2_DHW.tif")

sst_dhw <- terra::global(GBR_OISST2_DHW, mean, na.rm = TRUE) %>%
  mutate(date = as.Date(rownames(.))) |>
  mutate(
    year  = year(date),
    month = month(date),
    austral_year = if_else(month == 12L, year + 1L, year)  # Dec belongs to next year
  ) %>%
  #filter(month %in% c(12, 1, 2, 3)) %>%
  filter(austral_year > 1940, austral_year < 2026) %>%
  group_by(austral_year) %>%
  summarise(dhw = max(mean, na.rm = TRUE), .groups = "drop") %>%
  rename(year = austral_year)
 

sst_dhw <- sst_dhw %>% arrange(year)
y_dhw <- sst_dhw$dhw
t_dhw <- sst_dhw$year

beast_dhw <- beast(
  y_dhw,
  start  = t_dhw[1],
  deltat = 1,
  season = "none",
  print.progress = FALSE,
  quiet = TRUE
)

cp_dhw   <- beast_dhw$trend$cp
cppr_dhw <- beast_dhw$trend$cpPr
if (is.null(cppr_dhw)) cppr_dhw <- beast_dhw$trend$cpProb

sst_cp_tbl <- tibble(cp = cp_dhw, cp_prob = cppr_dhw) %>%
  mutate(year = as.integer(round(cp))) %>%
  arrange(desc(cp_prob))

breakpoints_dhw <- sst_cp_tbl$cp[1:min(2, nrow(sst_cp_tbl))]
breakyears_dhw  <- sst_cp_tbl$year[1:min(2, nrow(sst_cp_tbl))]



### fit models to BEAST CPD output
## SST
segment_1_sst <- sst_ts |> filter(year < (breakpoints_sst[1]))
segment_2_sst <- sst_ts |> filter(year > (breakpoints_sst[1]))
segment_mean_1_sst <- data.frame(year = c(min(segment_1_sst$year), max(segment_1_sst$year)), sst = rep(mean(segment_1_sst$sst),2))
segment_mean_2_sst <- data.frame(year = c(min(segment_2_sst$year), max(segment_2_sst$year)), sst = rep(mean(segment_2_sst$sst),2))

model_1_sst <- lm(sst ~ year, data = segment_1_sst)
model_2_sst <- lm(sst ~ year, data = segment_2_sst)

#summary(model_1_sst)
#summary(model_2_sst)

segment_1_sst_rect <- segment_1_sst |>
  summarise(
    start_year = min(year),
    end_year = max(year),
    ymin = min(sst),
    ymax = max(sst)
  )

segment_2_sst_rect <- segment_2_sst |>
  summarise(
    start_year = min(year),
    end_year = max(year),
    ymin = min(sst),
    ymax = max(sst)
  )

segment_1_sst_mean <- segment_1_sst |>
  summarise(
    year = mean(year),
    sst = mean(sst)
  )
segment_2_sst_mean <- segment_2_sst |>
  summarise(
    year = mean(year),
    sst = mean(sst)
  )


## SSTA

segment_1_ssta <- ssta_ts |> filter(year < (breakpoints_ssta[2]))
segment_2_ssta <- ssta_ts  |> filter(year > (breakpoints_ssta[2])) |> 
                              filter(year < (breakpoints_ssta[1]))
segment_3_ssta <- ssta_ts |> filter(year > (breakpoints_ssta[1]))
segment_mean_1_ssta <- data.frame(year = c(min(segment_1_ssta$year), max(segment_1_ssta$year)), ssta = rep(mean(segment_1_ssta$ssta),2))
segment_mean_2_ssta <- data.frame(year = c(min(segment_2_ssta$year), max(segment_2_ssta$year)), ssta = rep(mean(segment_2_ssta$ssta),2))
segment_mean_3_ssta <- data.frame(year = c(min(segment_3_ssta$year), max(segment_3_ssta$year)), ssta = rep(mean(segment_3_ssta$ssta),2))

model_1_ssta <- lm(ssta ~ year, data = segment_1_ssta)
model_2_ssta <- lm(ssta ~ year, data = segment_2_ssta)
model_3_ssta <- lm(ssta ~ year, data = segment_3_ssta)

#summary(model_1_ssta)
#summary(model_2_ssta)
#summary(model_3_ssta)

segment_1_ssta_rect <- segment_1_ssta |>
  summarise(
    start_year = min(year),
    end_year = max(year),
    ymin = min(ssta),
    ymax = max(ssta)
  )

segment_2_ssta_rect <- segment_2_ssta |>
  summarise(
    start_year = min(year),
    end_year = max(year),
    ymin = min(ssta),
    ymax = max(ssta)
  )

segment_3_ssta_rect <- segment_3_ssta |>
  summarise(
    start_year = min(year),
    end_year = max(year),
    ymin = min(ssta),
    ymax = max(ssta)
  )

segment_1_ssta_mean <- segment_1_ssta |>
  summarise(
    year = mean(year),
    ssta = mean(ssta)
  )
segment_2_ssta_mean <- segment_2_ssta |>
  summarise(
    year = mean(year),
    ssta = mean(ssta)
  )
segment_3_ssta_mean <- segment_3_ssta |>
  summarise(
    year = mean(year),
    ssta = mean(ssta)
  )



# plot main outputs

main_dhw <- ggplot() + theme_bw() +
 geom_vline(xintercept=c(1998, 2002, 2016, 2017, 2020, 2022, 2024, 2025), linewidth=2.8, alpha=0.1, color="coral4") +
 geom_col(data=sst_dhw, aes(x=year, y=dhw, fill=dhw), color="black", linewidth=0.25, show.legend=FALSE) +
 scale_fill_distiller(palette="YlOrRd", direction=1) + 
 #scale_fill_distiller(palette="RdBu", direction=-1) + 
 xlab("") + ylab("Degree Heating Weeks") +
 scale_y_continuous(limits=c(0,12), breaks=seq(0,10,2)) + 
 scale_x_continuous(limits=c(1940,2026), breaks=seq(1940,2025,10)) +
    theme(panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.grid.major.y = element_blank())  # Drop minor x grids) 


sst_break <- ggplot() + theme_bw() +
  geom_rect(data = segment_1_sst_rect, 
            aes(xmin = start_year, xmax = end_year, ymin = ymin, ymax = ymax), 
            fill="darkslategray4", alpha=0.1) + 
  geom_rect(data = segment_2_sst_rect, 
            aes(xmin = start_year, xmax = end_year, ymin = ymin, ymax = ymax), 
            fill="coral2", alpha=0.1) + 
  geom_line(data = segment_1_sst, 
            aes(x = year, y = sst), 
            color="darkgrey", linewidth=0.2) +
  geom_point(data = segment_1_sst, 
             aes(x = year, y = sst), 
             fill="darkslategray4", shape=21, size=1.5, show.legend=FALSE) +
  geom_line(data = segment_2_sst, 
            aes(x = year, y = sst), 
            color="darkgrey",linewidth=0.2) +
  geom_point(data = segment_2_sst, 
             aes(x = year, y = sst), 
             fill="coral2", shape=21, size=1.5, show.legend=FALSE) +
  scale_fill_distiller(palette="RdBu") +
  geom_vline(xintercept = breakpoints_sst[1], color = "darkgrey", linewidth=2, alpha=0.4) +
  geom_smooth(data = segment_2_sst, 
              aes(x = year, y = sst), 
             method = "lm", se = TRUE, linetype="solid", 
             color = "indianred4",  fill = "coral2", linewidth = 0.5, alpha=0.05) +
    geom_line(data=segment_mean_1_sst, aes(x=year, y=sst), linewidth=0.25, linetype="solid") +
  labs(x = "", y = "SST (°C)") +
  theme(
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    axis.text=element_text(size=6),
    axis.title=element_text(size=7),
    axis.ticks = element_line(linewidth = 0.25, colour = "darkgrey"),
    plot.background = element_rect(fill = "transparent", colour = NA),
    panel.border = element_rect(linewidth = 0, fill = "transparent"), 
    panel.grid.major.x = element_blank(), 
    panel.grid.minor.x = element_blank()) 


ssta_break <- ggplot() +
  theme_bw() +
  geom_rect(
    data = segment_1_ssta_rect,
    aes(xmin = start_year, xmax = end_year, ymin = ymin, ymax = ymax),
   fill="darkslategray4", alpha=0.1
  ) +
  geom_rect(data = segment_2_ssta_rect, 
            aes(xmin = start_year, xmax = end_year, ymin = ymin, ymax = ymax), 
            fill = "coral2", alpha = 0.1) +
  geom_rect(
    data = segment_3_ssta_rect,
    aes(xmin = start_year, xmax = end_year, ymin = ymin, ymax = ymax),
    fill = "indianred4", alpha = 0.1
  ) +
  geom_line(data = segment_1_ssta, aes(x = year, y = ssta), color = "darkgrey", linewidth = 0.2) +
  geom_point(data = segment_1_ssta, aes(x = year, y = ssta), fill = "darkslategray3", shape = 21, alpha = 1, size = 1.5, show.legend = FALSE) +
  geom_line(data = segment_2_ssta, aes(x = year, y = ssta), color = "darkgrey", linewidth = 0.2) +
  geom_point(data = segment_2_ssta, aes(x = year, y = ssta), fill = "indianred1", shape = 21, alpha = 1, size = 1.5, show.legend = FALSE) +
  geom_line(data = segment_3_ssta, aes(x = year, y = ssta), color = "darkgrey", linewidth = 0.2) +
  geom_point(data = segment_3_ssta, aes(x = year, y = ssta), fill = "indianred4", shape = 21, alpha = 1, size = 1.5, show.legend = FALSE) +
  scale_fill_distiller(palette = "RdBu") +
  scale_x_continuous(limits = c(1940, 2026)) +
  scale_y_continuous(limits = c(-0.6, 1.2)) +
  geom_vline(xintercept = breakpoints_ssta[1], color = "darkgrey", linewidth=2, alpha=0.4) +
  geom_vline(xintercept = breakpoints_ssta[2], color = "darkgrey", linewidth=2, alpha=0.4) +
  geom_line(data=segment_mean_1_ssta, aes(x=year, y=ssta), linewidth=0.25, linetype="solid") +
  geom_line(data=segment_mean_2_ssta, aes(x=year, y=ssta), linewidth=0.25, linetype="solid") +
  geom_line(data=segment_mean_3_ssta, aes(x=year, y=ssta), linewidth=0.25, linetype="solid") +
  labs(x = "", y = "SSTA (°C)") +# scale_y_continuous(limits=c(-0.5, 1.5)) +
  theme(
    axis.text=element_text(size=6),
    axis.title=element_text(size=7),
    axis.ticks = element_line(linewidth = 0.25, colour = "darkgrey"),
    plot.background = element_rect(fill = "transparent", colour = NA),
    panel.border = element_rect(linewidth = 0, fill = "transparent"), # Change border thickness
    panel.grid.minor.y = element_blank(), # Drop minor y grids
    panel.grid.major.x = element_blank(), # Drop major x grids
    panel.grid.minor.x = element_blank(),
    plot.margin = margin(t = 10, r = 10, b = 0, l = 10)) 



inset <- sst_break / ssta_break + theme(plot.margin = margin(-1, 1, 0, 0))

df.labels <- data.frame(date=c(1998, 2002, 2016, 2017, 2020, 2022, 2024, 2025),
                        label=c(1998, 2002, 2016, 2017, 2020, 2022, 2024, 2025), 
                        dhw = rep(11.8, 8))

```

```{r}
#| message: false
#| warning: false

######################################################################-
# Rate of change positive SST Anomlies 1981-2024 
######################################################################-

# load OISST
library(sf)


# Define the function to calculate the slope across the entire time series
slope <- terra::app(GBR_OISST_SSTA, function(sst_ts) {
  if (sum(!is.na(sst_ts)) > 1) {  # Ensure there are more than one non-NA values
    df <- data.frame(sst = sst_ts, time = seq_along(sst_ts))  # Create a time series index
    lm_fit <- stats::lm(sst ~ time, data = df)  # Fit linear model (sst as response, time as predictor)
    return(coef(lm_fit)[2])  # Return the slope (the coefficient for time)
  } else {
    return(NA)  # Return NA if not enough data
  }
})

annual_rate_of_change <- slope * 365.25 *10

# Create grid cells covering the raster extent
grid_cells <- st_bbox(c(xmin = xmin(annual_rate_of_change),
  ymin = ymin(annual_rate_of_change),
  xmax = xmax(annual_rate_of_change),
  ymax = ymax(annual_rate_of_change)),
  crs = st_crs(4326)) |>
st_as_sfc() |>
st_make_grid(cellsize = res(GBR_OISST_SSTA)[1], what = "polygons")
#grid_cells <- st_sf(id = 1:length(grid_cells), geometry = grid_cells)


# Create a polygon of non-NA cells in GBR_OISST_SSTA
mask <- mask(annual_rate_of_change, annual_rate_of_change >= 0.05)
mask[!is.na(mask)] <- 1

mask_points <- as.points(mask, na.rm = TRUE)
mask_sf <- st_as_sf(mask_points)
projected_crs <- "+proj=utm +zone=55 +south +datum=WGS84 +units=m +no_defs"
mask_projected <- st_transform(mask_sf, crs = projected_crs)
resolution_meters <- 25000  # 0.25 
mask_buffered <- mask_projected |>
  st_buffer(dist = resolution_meters/2, endCapStyle = "SQUARE")
mask_polygons <- st_transform(mask_buffered, crs = st_crs(mask_sf))


# Define the numeric breaks for the values and color scale
anom_numeric_breaks <- c(0.1, 0.15, 0.2, 0.25, 0.3)  # Numeric values to map colors
anom_colour_breaks <- c("#ffffff", "#f4a6a6", "#e03f3f", "#680202")  # Colors to use at these breaks

rate_change_plot <- ggplot() + 
  theme_bw() + 
  tidyterra::geom_spatraster(data = annual_rate_of_change,  interpolate = FALSE, show.legend = TRUE) +
  scale_fill_gradientn(
    limits = c(0.1, 0.3),
    name = "SST anom",
    na.value = "transparent",
    breaks = anom_numeric_breaks,       # Set custom breaks for the color scale
    labels = anom_numeric_breaks,       # Use the same values for labels
    colours = anom_colour_breaks,       # Apply colors directly
    values = scales::rescale(anom_numeric_breaks, from = c(0.1, 0.3))  # Rescale numeric breaks to match color scale
  ) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1),
    plot.title = element_text(size = 7, hjust = 0.5),
    legend.position = c(0.95, 0.95), 
    legend.justification = c(1, 1),  
    legend.text = element_text(size = 4, margin = margin(t = 0, r = 1, b = 0, l = 1.5)),  # Reduce margin around text
    legend.title = element_text(size = 5),  # Scale legend title
    legend.key.size = unit(0.5, "lines"),  # Scale legend box size
    legend.background = element_rect(fill = "white", color = NA), 
    legend.margin = margin(2.1, 2.1, 2.1, 2.1)
  ) +
  geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1) + 
  geom_sf(data = ausmap, fill = "#E2E7D4", alpha = 1) +
  labs(title = "Decadal Rate of Change \nin SST anom (°C) 1982-2024", fill = "SST") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))



######################################################################-
# Maximum Degree Heating Week (1981-2025)
######################################################################-


# Calculate the maximum DHW across all layers
max_DHW <- max(GBR_OISST_DHW, na.rm = TRUE)

# Define the numeric breaks for the values
dhw_numeric_breaks <- c(4, 8, 12, 16, 20)  # Numeric values to map colors
#dhw_colour_breaks <- c(, "#eddf8c", "#ff9a15", "#820000")  # Colors to use at these breaks
dhw_colour_breaks <- c( "#a5d3e5", "#eddf8c",  "#d47a04",  "#4d0101")  # Colors to use at these breaks



# Plot
max_DHW_plot <- ggplot() + 
  theme_bw() + 
  ggtitle("Max DHW 1982-2025") +
  tidyterra::geom_spatraster(data = max_DHW, interpolate = FALSE, show.legend = TRUE) + 
  scale_fill_gradientn(
    limits = c(4, 20),
    name = "Max DHW",
    na.value = "transparent",
    breaks = dhw_numeric_breaks,       # Set custom breaks for the color scale
    labels = dhw_numeric_breaks,       # Use the same values for labels
    colours = dhw_colour_breaks,       # Apply colors directly
    values = scales::rescale(dhw_numeric_breaks, from = c(4, 20))  # Rescale numeric breaks to match color scale
  ) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1),
    plot.title = element_text(size = 7, hjust = 0.5),
    axis.text.y = element_blank(),
    legend.position = c(0.95, 0.95),
    legend.justification = c(1, 1),  
    legend.text = element_text(size = 4, margin = margin(t = 0, r = 1, b = 0, l = 1.5)),
    legend.title = element_text(size = 5),
    legend.key.size = unit(0.5, "lines"),
    legend.background = element_rect(fill = "white", color = NA), 
    legend.margin = margin(2.1, 2.1, 2.1, 2.1)
  ) +
  geom_sf(data = ausmap, fill = "#E2E7D4", alpha = 1) +
  labs(title = "Maximum DHW \nexposure (1982-2025)", fill = "DHW") +
  geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1)  

######################################################################-
# Warmest year (DHW)
######################################################################-


# Extract the time information (days) and convert to years
time_info <- time(GBR_OISST_DHW)
years <- as.numeric(format(time_info, "%Y"))

# Aggregate daily data to annual maximum DHW
maxDHW_annual <- tapp(GBR_OISST_DHW, index = years, fun = max, na.rm = TRUE)

df <- as.data.frame(maxDHW_annual, xy = TRUE)
colnames(df) <- gsub("^X", "", colnames(df))

df_long <- df %>%
  pivot_longer(cols = -c(x,y), 
               names_to = "year", 
               values_to = "dhw") 

df_long_max <- df_long %>%
  group_by(x, y) %>%
  mutate(max_dhw = max(dhw, na.rm = TRUE),  # Find the max DHW for each group
         dhw = ifelse(dhw == max_dhw, dhw, NA),  # Set non-max dhw values to NA
         year = ifelse(!is.na(dhw), year, NA)) %>%  # Set non-max year values to NA
  dplyr::select(-max_dhw) |>   # Remove the temporary max_dhw column
  drop_na(year) |> 
  arrange(year)

df_wide_max <- df_long_max %>%
  pivot_wider(names_from = year, values_from = dhw) %>%
  mutate(across(`1982`:`2025`, ~ as.factor(ifelse(!is.na(.), cur_column(), NA))))

year_max <- rast(df_wide_max, type = "xyz")

# Use the `app` function to merge the layers, keeping the first non-NA value
flattened_raster <- app(year_max, fun = function(x) {
  first_non_na <- x[!is.na(x)]
  if (length(first_non_na) > 0) {
    return(first_non_na[1])  # Return the first non-NA value
  } else {
    return(0)  # If all values are NA, return NA
  }
})

values(flattened_raster) <- as.factor(values(flattened_raster))
flattened_raster <- droplevels(flattened_raster) # Drop NA factor levels
# Set the CRS if needed
crs(flattened_raster) <- "EPSG:4326"


library(RColorBrewer)
years <- as.character(1982:2025)
n_colors <- length(years)
#spectral_palette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))(n_colors)
spectral_palette <- colorRampPalette(c(rev(brewer.pal(6, "Greens")), rev(brewer.pal(8, "Blues")), (brewer.pal(8, "Reds"))))(n_colors)
year_colors <- setNames(spectral_palette, years)
# 
# max_dhw_year_plot <- ggplot() + theme_bw() + 
#   ggtitle("") +
#   tidyterra::geom_spatraster(data=flattened_raster, interpolate=FALSE, show.legend=TRUE) +
#   scale_fill_manual(values = year_colors, na.value = "transparent", guide = guide_legend(na.translate = FALSE)) +
#   theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=1)) +
#   geom_sf(data=ausmap, fill="#E2E7D4", alpha=1) + 
#   labs(title="Maximum DHW \nexposure (year)", fill="year") +
#   theme(plot.title = element_text(size=6, hjust = 0.5),
#         axis.text.y = element_blank(),
#         legend.position = c(0.95, 0.95), 
#         legend.justification = c(1, 1),  
#         legend.text = element_text(size = 3, margin = margin(t = 0, r = 1, b = 0, l = 1.5)),  # Reduce margin around text
#         legend.title = element_text(size = 4),  # Scale legend title to 35%
#         legend.key.size = unit(0.25, "lines"),  # Scale legend box size to 35%
#         legend.background = element_rect(fill = "white", color = NA), 
#         legend.margin = margin(1,1,1,1)) +
#   geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1)  
# 

######################################################################- 
# Frequency of Degree Heating Weeks exceeding 6DHW
######################################################################-

dhw_year_max <- GBR_OISST_DHW %>%
  tapp(index = format(time(.), "%Y"), fun = max, na.rm = TRUE)


# Create hull polygon
hull <- data.frame(lat = c(-9, -25, -25), 
                   lon = c(136, 136, 151.5)) |> 
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326) |> 
  st_combine() |>            
  st_convex_hull() |>        
  st_cast("POLYGON")  

# Create outline polygon and remove intersection with hull
ausoutlinefill <- ausmap |> 
  st_cast("MULTILINESTRING") |> 
  st_cast("POLYGON") |> 
  dplyr::select(scalerank) |>  
  st_difference(hull)  # Get parts outside the hull


ausoutline <- ausmap |>
  st_cast("MULTILINESTRING") |>
  dplyr::select(scalerank) |>
  st_crop(st_as_sfc(st_bbox(c(xmin = 142.05, ymin = -24.75, xmax = 154.00, ymax = -8.75), crs = 4326))) |>
  st_cast("LINESTRING") |>
  mutate(length_km = st_length(geometry) / 1000) |>
  arrange(desc(length_km))

ausoutline <- ausoutline[1,]

######################################################################- 
# Frequency of Degree Heating Weeks (1986-1995)
######################################################################-

#dhw_subset_first <- subset(dhw_year_max, layer_names)
dhw_subset_first <- subset(dhw_year_max, paste0("X", seq(1986, 1995, 1)))
dhw_exceeding_6_first_subset <- app(dhw_subset_first, fun = function(x) {
  sum(x > 6, na.rm = TRUE)
})
dhw_exceeding_6_first_subset <- subst(dhw_exceeding_6_first_subset, from = 0, to = NA)
values(dhw_exceeding_6_first_subset) <- values(dhw_exceeding_6_first_subset) |> as.factor()
# 
# 
# 
# dhw_exceedance_1985_1994 <- ggplot() + theme_bw() + 
#   tidyterra::geom_spatraster(data=dhw_exceeding_6_first_subset, interpolate=FALSE, show.legend=FALSE) +
#   #scale_fill_distiller(palette="Spectral", na.value="transparent", breaks=seq(0,7,1), limits = c(0, 7)) +
#   scale_fill_manual(values = colorRampPalette(rev(brewer.pal(11, "Spectral")))(7), 
#                   na.value = "transparent",
#                   guide = guide_legend(na.translate = FALSE)) +
#   theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=1)) +
#   labs (title="Bleaching threshold exceedance \nfrequency 1985-1995", fill="n \nevents", ylab=NULL) +
#   geom_sf(data=ausmap, fill="#E2E7D4", alpha=1) +
#    theme(plot.title = element_text(size=6, hjust = 0.5),
#         axis.text.y = element_blank(),
#         legend.position = c(0.95, 0.95), 
#         legend.justification = c(1, 1),  
#         legend.text = element_text(size = 3, margin = margin(t = 0, r = 1, b = 0, l = 1)),  # Reduce margin around text
#         legend.title = element_text(size = 4),  # Scale legend title to 35%
#         legend.key.size = unit(0.25, "lines"),  # Scale legend box size to 35%
#         legend.background = element_rect(fill = "white", color = NA), 
#         legend.margin = margin(0.1,0.1,0.1,0.1))  +
#   geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1)  
# 


######################################################################- 
# Frequency of Degree Heating Weeks (1996-2005)
######################################################################-


dhw_subset_second <- subset(dhw_year_max, paste0("X", seq(1996, 2005, 1)))
dhw_exceeding_6_second_subset <- app(dhw_subset_second, fun = function(x) {
  sum(x > 6, na.rm = TRUE)
})
dhw_exceeding_6_second_subset <- subst(dhw_exceeding_6_second_subset, from = 0, to = NA)
values(dhw_exceeding_6_second_subset) <- values(dhw_exceeding_6_second_subset) |> as.factor()

# 
# dhw_exceedance_1995_2004 <- ggplot() + theme_bw() + 
#   tidyterra::geom_spatraster(data=dhw_exceeding_6_second_subset, interpolate=FALSE, show.legend=FALSE) +
#   #scale_fill_distiller(palette="Spectral", na.value="transparent", breaks=seq(0,7,1), limits = c(0, 7)) +
#   scale_fill_manual(values = colorRampPalette(rev(brewer.pal(11, "Spectral")))(7), 
#                   na.value = "transparent",
#                   guide = guide_legend(na.translate = FALSE)) +
#   theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=1)) +
#   labs (title="Bleaching threshold exceedance \nfrequency 1995-2005", fill="n \nevents", ylab=NULL) +
#   geom_sf(data=ausmap, fill="#E2E7D4", alpha=1) +
#    theme(plot.title = element_text(size=6, hjust = 0.5),
#         axis.text.y = element_blank(),
#         legend.position = c(0.95, 0.95), 
#         legend.justification = c(1, 1),  
#         legend.text = element_text(size = 3, margin = margin(t = 0, r = 1, b = 0, l = 1)),  # Reduce margin around text
#         legend.title = element_text(size = 4),  # Scale legend title to 35%
#         legend.key.size = unit(0.25, "lines"),  # Scale legend box size to 35%
#         legend.background = element_rect(fill = "white", color = NA), 
#         legend.margin = margin(0.1,0.1,0.1,0.1))  +
#   geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1)  

######################################################################- 
# Frequency of Degree Heating Weeks (2006-2015)
######################################################################-


dhw_subset_third <- subset(dhw_year_max, paste0("X", seq(2006, 2015, 1)))
dhw_exceeding_6_third_subset <- app(dhw_subset_third, fun = function(x) {
  sum(x > 6, na.rm = TRUE)
})
dhw_exceeding_6_third_subset <- subst(dhw_exceeding_6_third_subset, from = 0, to = NA)
values(dhw_exceeding_6_third_subset) <- values(dhw_exceeding_6_third_subset) |> as.factor()
# 
# dhw_exceedance_2005_2014 <- ggplot() + theme_bw() + 
#   tidyterra::geom_spatraster(data=dhw_exceeding_6_third_subset, interpolate=FALSE, show.legend=FALSE) +
#   #scale_fill_distiller(palette="Spectral", na.value="transparent", breaks=seq(0,7,1), limits = c(0, 7)) +
#   scale_fill_manual(values = colorRampPalette(rev(brewer.pal(11, "Spectral")))(7), 
#                   na.value = "transparent",
#                   guide = guide_legend(na.translate = FALSE)) +
#   theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=1)) +
#   labs (title="Bleaching threshold exceedance \nfrequency 2005-2015", fill="n \nevents", ylab=NULL) +
#   geom_sf(data=ausmap, fill="#E2E7D4", alpha=1) +
#    theme(plot.title = element_text(size=6, hjust = 0.5),
#         axis.text.y = element_blank(),
#         legend.position = c(0.95, 0.95), 
#         legend.justification = c(1, 1),  
#         legend.text = element_text(size = 3, margin = margin(t = 0, r = 1, b = 0, l = 1)),  # Reduce margin around text
#         legend.title = element_text(size = 4),  # Scale legend title to 35%
#         legend.key.size = unit(0.25, "lines"),  # Scale legend box size to 35%
#         legend.background = element_rect(fill = "white", color = NA), 
#         legend.margin = margin(0.1,0.1,0.1,0.1))  +
#   geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1)  
# 

######################################################################- 
# Frequency of Degree Heating Weeks (2016-2025)
######################################################################-

dhw_subset_last <- subset(dhw_year_max, paste0("X", seq(2016, 2025, 1)))
dhw_exceeding_6_subset_last <- app(dhw_subset_last, fun = function(x) {
  sum(x > 6, na.rm = TRUE)
})
dhw_exceeding_6_subset_last <- subst(dhw_exceeding_6_subset_last, from = 0, to = NA)
values(dhw_exceeding_6_subset_last) <- values(dhw_exceeding_6_subset_last) |> as.factor()

# 
# 
# dhw_exceedance_2015_2024 <- ggplot() + theme_bw() + 
#   tidyterra::geom_spatraster(data=dhw_exceeding_6_subset_last, interpolate=FALSE) +
#   #scale_fill_distiller(palette="Spectral", na.value="transparent", breaks=seq(0,7,1), limits = c(0, 7)) +
#   scale_fill_manual(values = colorRampPalette(rev(brewer.pal(11, "Spectral")))(7), 
#                   na.value = "transparent",
#                   guide = guide_legend(na.translate = FALSE)) +
#   geom_sf(data=ausmap, fill="#E2E7D4", alpha=1) + 
#   labs(title="Bleaching threshold exceedance \nfrequency 2015-2025", ylab = "", fill="n \nevents") + 
#   theme(plot.title = element_text(size=6, hjust = 0.5),
#         axis.text.x = element_text(angle = 90, vjust = 1, hjust=1, size=8),
#         axis.text.y = element_blank(),
#         axis.title.y = element_blank(),
#         legend.position = c(0.95, 0.95), 
#         legend.justification = c(1, 1),  
#         legend.text = element_text(size = 4, margin = margin(t = 0, r = 1, b = 0, l = 1.5)),  # Reduce margin around text
#         legend.title = element_text(size = 5),  # Scale legend title to 35%
#         legend.key.size = unit(0.3, "lines"),  # Scale legend box size to 35%
#         legend.background = element_rect(fill = "white", color = NA), 
#         legend.margin = margin(0.1,0.1,0.1,0.1))  +
#   geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1)  




# hack to set extent with lat
raster_df <- as.data.frame(annual_rate_of_change, xy = TRUE, na.rm = TRUE)
current_extent <- ext(annual_rate_of_change)
longitude_range <- xmax(current_extent) - xmin(current_extent)
extended_xmax <- xmax(current_extent) +  longitude_range*1.5  # Extend 3x the longitude range

mask_polygons_border <- as.polygons(mask, dissolve = TRUE)
mask_sf <- st_as_sf(mask_polygons_border)
mask_border <- st_union(mask_sf)






GBR_OISST2_SST <- rast("~/GBR-dhw/datasets/GBR_OISST2_SST.tif")


# # maxSST per year for OISST2 in rast
gbr_OISST2_dhw_annual_max <- summarise_raster(GBR_OISST2_DHW, index = "years", fun = max, na.rm = TRUE, overwrite=TRUE)



# hack to set extent with lat
raster_df <- as.data.frame(annual_rate_of_change, xy = TRUE, na.rm = TRUE)
current_extent <- ext(annual_rate_of_change)
longitude_range <- xmax(current_extent) - xmin(current_extent)
extended_xmax <- xmax(current_extent) +  longitude_range  # Extend 3x the longitude range

mask_polygons_border <- as.polygons(mask, dissolve = TRUE)
mask_sf <- st_as_sf(mask_polygons_border)
mask_border <- st_union(mask_sf)


################## 1. rate change SSTA ########################-
rate_change_plot <- ggplot() + 
  theme_bw() +
  
  # Y-axis with exactly 6 evenly spaced labels
 scale_y_continuous(limits = c(-25, -9),
                     breaks=seq(-25, -9, 2),
                     labels = paste0(seq(-25, -9, 2), "°N"))  +

  # X-axis with rotated labels
  scale_x_continuous(
    limits = c(xmin(current_extent), extended_xmax), 
    breaks = seq(140, 152, 2),
    labels = paste0(seq(140, 152, 2), "°E")
  ) +

  # Plot components
  tidyterra::geom_spatraster(data = annual_rate_of_change, interpolate = FALSE, show.legend = TRUE) +
  scale_fill_gradientn(
    guide = guide_colorbar(
      frame.colour = "black",
      frame.linewidth = 0.2,
      ticks.colour = "black",
      ticks.linewidth = 0.2
    ),
    limits = c(0.175, 0.3),
    name = "SSTA",
    na.value = "transparent",
    breaks = anom_numeric_breaks,
    labels = anom_numeric_breaks,
    colours = anom_colour_breaks,
    values = scales::rescale(anom_numeric_breaks, from = c(0.1, 0.3))
  ) +

  # Theme settings
  theme(
    title = NULL,
    axis.text.x = element_text(size = 6, angle = 90, vjust = 0.5, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 6),
    legend.position = c(0.28, 0.97),
    legend.justification = c(1, 1),
    legend.text = element_text(size = 5, margin = margin(t = 0, r = 1, b = 0, l = 3.5)),
    legend.title = element_text(size = 6),
    legend.key.size = unit(0.5, "lines"),
    legend.background = element_rect(fill = "transparent", color = NA), 
    legend.margin = margin(3.1, 3.1, 3.1, 3.1),
    legend.key = element_rect(fill = "white", color = "black")
  ) +

  # Overlay other geometries
  geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1) + 
  geom_sf(data = mask_border, fill = NA, linewidth = 0.5, alpha = 1) +
  geom_sf(data = ausmap, fill = "#E2E7D4", alpha = 1) +
  
  # Label settings
  labs(fill = "SST")



################## 2. max DHW ########################-

max_DHW_plot <- ggplot() + 
  theme_void() +
  
  # Spatial raster layer
  tidyterra::geom_spatraster(data = max_DHW, interpolate = FALSE, show.legend = TRUE) +
  
  # Color scale for DHW
  scale_fill_gradientn(
    guide = guide_colorbar(
      frame.colour = "black",
      frame.linewidth = 0.2,
      ticks.colour = "black",
      ticks.linewidth = 0.2
    ),
    limits = c(4, 20),
    name = bquote(DHW[max]),
    na.value = "transparent",
    breaks = dhw_numeric_breaks,
    labels = dhw_numeric_breaks,
    colours = dhw_colour_breaks,
    values = scales::rescale(dhw_numeric_breaks, from = c(4, 20))
  ) +

  # Theme and legend styling
  theme(
    title = NULL,
    axis.text.x = NULL,
    legend.position = c(0.5, 0.97),
    legend.justification = c(1, 1),
    legend.text = element_text(size = 5, margin = margin(t = 0, r = 1, b = 0, l = 3.5)),
    legend.title = element_text(size = 6),
    legend.key.size = unit(0.5, "lines"),
    legend.background = element_rect(fill = "transparent", color = NA),
    legend.margin = margin(3.1, 3.1, 3.1, 3.1),
    legend.key = element_rect(fill = "white", color = "black")
  ) +
  
  # Spatial data layers
  geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1) +
  geom_sf(data = mask_border, fill = NA, linewidth = 0.5, alpha = 1) +
  
  # Labels
  labs(fill = "DHW")



################## 3. max DHW ##################-

max_dhw_year_plot <- ggplot() + 
  
  theme_void() +
  
  # Spatial raster layer
  tidyterra::geom_spatraster(data = flattened_raster, interpolate = FALSE, show.legend = TRUE) +

  # Manual color scale for year
  scale_fill_manual(
    values = year_colors,
    na.value = "transparent",
    guide = guide_legend(na.translate = FALSE)
  ) +
  
  # Theme and legend styling
  theme(
    title = NULL,
    axis.text.x =NULL,
    legend.position = c(0.78, 0.98),
    legend.justification = c(1, 1),
    legend.text = element_text(size = 5, margin = margin(t = 0, r = 1, b = 0, l = 3.5)),
    legend.title = element_text(size = 6),
    legend.key.size = unit(0.4, "lines"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.margin = margin(3.1, 3.1, 3.1, 3.1),
    legend.key = element_rect(fill = "white", color = "black")
  ) +
  
  # Legend layout (2 columns)
  guides(fill = guide_legend(ncol = 2)) +

  # Spatial data layers
  geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1) +
  geom_sf(data = mask_border, fill = NA, linewidth = 0.5, alpha = 1) +

  # Labels
  labs(title = "", fill = "Year")







################## 1. dhw_exceedance_1985_1994 ##################-



dhw_exceedance_1985_1994 <- ggplot() +
  theme_bw() +
  scale_y_continuous(limits = c(-25, -9),
                     breaks=seq(-25, -9, 2),
                     labels = paste0(seq(-25, -9, 2), "°N"))  +
  scale_x_continuous(limits = c(xmin(current_extent), extended_xmax),
                     breaks=seq(140, 152, 2),
                     labels = paste0(seq(140, 152, 2), "°E"))  +
  tidyterra::geom_spatraster(data = dhw_exceeding_6_first_subset,  interpolate = FALSE, show.legend = FALSE) +
  scale_fill_manual(values = colorRampPalette(rev(brewer.pal(11, "Spectral")))(7),
                    na.value = "transparent"
                    #guide = guide_legend(na.translate = FALSE),
                  ) +
  theme(
    title = NULL,
    axis.text.x = element_text(size = 6, angle = 90, vjust = 0.5, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 6),
    legend.position = c(0.75, 0.98),
    legend.justification = c(1, 1),
    legend.text = element_text(size = 5, margin = margin(t = 0, r = 1, b = 0, l = 3.5)),
    legend.title = element_text(size = 7),
    legend.key.size = unit(0.75, "lines"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.margin = margin(3.1, 3.1, 3.1, 3.1),
    legend.key = element_rect(fill = "white", color = "black")  # Add black border around legend keys
  ) +
  geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1) +
  geom_sf(data = mask_border, fill = NA, linewidth=0.5, alpha = 1) +
  geom_sf(data = ausmap, fill = "#E2E7D4", alpha = 1) +
  labs(fill = "n")

################## 2. dhw_exceedance_1995_2004 ##################-

dhw_exceedance_1995_2004 <- ggplot() +
  theme_void() +
  scale_y_continuous(limits = c(-25, -9),
                     breaks=seq(-25, -9, 2),
                     labels = paste0(seq(-25, -9, 2), "°N"))  +

  tidyterra::geom_spatraster(data = dhw_exceeding_6_second_subset,  interpolate = FALSE, show.legend = FALSE) +
  scale_fill_manual(values = colorRampPalette(rev(brewer.pal(11, "Spectral")))(7),
                    na.value = "transparent"
                    #guide = guide_legend(na.translate = FALSE),
                  ) +
  theme(
    title = NULL,
    axis.text.x = NULL,
    legend.position = c(0.56, 0.98),
    legend.justification = c(1, 1),
    legend.text = element_text(size = 5, margin = margin(t = 0, r = 1, b = 0, l = 3.5)),
    legend.title = element_text(size = 7),
    legend.key.size = unit(0.75, "lines"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.margin = margin(3.1, 3.1, 3.1, 3.1),
    legend.key = element_rect(fill = "white", color = "black")  # Add black border around legend keys
  ) +
  geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1) +
  geom_sf(data = mask_border, fill = NA, linewidth=0.5, alpha = 1) +
 # geom_sf(data = ausmap, fill = "#E2E7D4", alpha = 1) +
  labs(fill = "n")


################## 3. dhw_exceedance_2005_2014 ##################-

dhw_exceedance_2005_2014 <- ggplot() +
  theme_void() +
  scale_y_continuous(limits = c(-25, -9),
                     breaks=seq(-25, -9, 2),
                     labels = paste0(seq(-25, -9, 2), "°N"))  +

  tidyterra::geom_spatraster(data = dhw_exceeding_6_third_subset,  interpolate = FALSE, show.legend = FALSE) +
  scale_fill_manual(values = colorRampPalette(rev(brewer.pal(11, "Spectral")))(7),
                    na.value = "transparent"
                    #guide = guide_legend(na.translate = FALSE),
                  ) +
  theme(
    title = NULL,
    axis.text.x = NULL,
    legend.position = c(0.75, 0.98),
    legend.justification = c(1, 1),
    legend.text = element_text(size = 5, margin = margin(t = 0, r = 1, b = 0, l = 3.5)),
    legend.title = element_text(size = 6),
    legend.key.size = unit(0.75, "lines"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.margin = margin(3.1, 3.1, 3.1, 3.1),
    legend.key = element_rect(fill = "white", color = "black")  # Add black border around legend keys
  ) +
  geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1) +
  geom_sf(data = mask_border, fill = NA, linewidth=0.5, alpha = 1) +
  #geom_sf(data = ausmap, fill = "#E2E7D4", alpha = 1) +
  labs(fill = "n")

dhw_exceeding_6_subset_last_2 <- dhw_exceeding_6_subset_last
dhw_exceeding_6_subset_last_2 <- classify(dhw_exceeding_6_subset_last_2, cbind(NA, 0))
dhw_exceeding_6_subset_last_2 <- as.factor(dhw_exceeding_6_subset_last_2)


################## 4. dhw_exceedance_2015_2024 ##################-

dhw_exceedance_2015_2024 <- ggplot() +
  theme_void() +
  scale_y_continuous(limits = c(-25, -9),
                     breaks=seq(-25, -9, 2),
                     labels = paste0(seq(-25, -9, 2), "°N"))  +
  tidyterra::geom_spatraster(data = dhw_exceeding_6_subset_last_2,  interpolate = FALSE, show.legend = TRUE) +
   scale_fill_manual(
    values = c("transparent", colorRampPalette(rev(brewer.pal(11, "Spectral")))(7)),  # Adjust color for 0-7 (8 levels)
    breaks = 0:7,                           # Order legend from 0 to 7
    labels = as.character(0:7),             # Labels from 0 to 7
    na.value = "transparent",               # Ensure NA is not visualized
    name = "> 6DHW events \n per decade"                       # Legend title
  ) +
  theme(
    title = NULL,
    axis.text.x = NULL,
    legend.position = c(0.85, 0.98),
    legend.justification = c(1, 1),
    legend.text = element_text(size = 5, margin = margin(t = 0, r = 1, b = 0, l = 3.5)),
    legend.title = element_text(size = 6.5, hjust=0.2),
    legend.key.size = unit(0.4, "lines"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.margin = margin(3.1, 3.1, 3.1, 3.1),
    legend.key = element_rect(fill = "white", color = "black")  # Add black border around legend keys
  ) +
  geom_sf(data = mask_polygons, fill = NA, linewidth = 0.1) +
  geom_sf(data = mask_border, fill = NA, linewidth=0.5, alpha = 1) +
  #geom_sf(data = ausmap, fill = "#E2E7D4", alpha = 1) +
  labs(fill = "N events")

library(patchwork)

dhw_6dhw_maps <- dhw_exceedance_1985_1994 +
  inset_element(dhw_exceedance_1995_2004, left = -0.095, bottom = 0, right = 0.905, top = 1) +
  inset_element(dhw_exceedance_2005_2014, left = 0.1, bottom = 0, right = 1.1, top = 1) +
  inset_element(dhw_exceedance_2015_2024, left = 0.28, bottom = 0, right = 1.28, top = 1)


######## COMBINED-

mainfig1 <- main_dhw +
  geom_text(data=df.labels, aes(date, dhw, label=label), size=2.5, angle = 90) +
  inset_element(inset, left = 0.02, bottom = 0.3, right = 0.58, top = 0.98)  

dhw_maps <- rate_change_plot +
  inset_element(max_DHW_plot, left = 0.01, bottom = 0, right = 1.01, top = 1) +
  inset_element(max_dhw_year_plot, left = 0.245, bottom = 0, right = 1.245, top = 1.04)

```

## Supplemenentary Figures
### Figure S1 Tipping points in DHW

Annual Great Barrier Reef-wide sea surface temperature (SST), sea surface temperature anomaly (SSTA), and Degree Heating Weeks (`DHW`) from 1940–2025 were analysed using Bayesian changepoint detection  (`beast`). Columns show `SST`, `SSTA` and `DHW`, respectively. Rows show the fitted trend component with observations and posterior mean trend in red, posterior changepoint probability `Pr(tcp)`, posterior probability of trend order, posterior slope-sign probabilities, and residual error. Dashed vertical lines indicate the most probable changepoint years. `SST` and `SSTA` show a marked shift from weak or near-stationary trends before the late 1990s to sustained warming thereafter, while DHW shows increasingly frequent and intense thermal-stress peaks after 1998, with the strongest events concentrated in the most recent decade.

```{r}
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 12

gbr_OISST2_dhw_annual_max <- summarise_raster(GBR_OISST2_DHW, index = "years", fun = max, na.rm = TRUE, overwrite=TRUE)


library(ggplot2)
library(patchwork)
library(tibble)
library(dplyr)
library(tidyr)

ggbeast <- function(o, title = NULL, point_size = 1.6, line_width = 0.8, top_n_vlines = 2) {
  stopifnot(inherits(o, "beast"))
  
  df <- tibble(
    time   = as.numeric(o$time),
    data   = as.numeric(o$data),
    trend  = as.numeric(o$trend$Y),
    tcpPr  = as.numeric(o$trend$cpOccPr),
    orderT = as.numeric(o$trend$order),
    err    = as.numeric(o$data) - as.numeric(o$trend$Y)
  )
  
  slp_df <- tibble(
    time = df$time,
    pos  = as.numeric(o$trend$slpSgnPosPr),
    zero = as.numeric(o$trend$slpSgnZeroPr)
  ) %>%
    mutate(neg = pmax(0, 1 - pos - zero)) %>%
    pivot_longer(cols = c(pos, zero, neg), names_to = "sign", values_to = "prob") %>%
    mutate(sign = factor(sign, levels = c("neg", "zero", "pos")))
  
  # vlines from BEAST trend cp + cpPr
  cp_tbl <- tibble(
    cp = as.numeric(o$trend$cp),
    cp_prob = as.numeric(o$trend$cpPr)
  ) %>%
    mutate(year = as.integer(round(cp))) %>%
    arrange(desc(cp_prob))
  
  vlines <- cp_tbl$cp[1:min(top_n_vlines, nrow(cp_tbl))]
  
  vline_layer <- if (length(vlines) == 0) NULL else
    geom_vline(xintercept = vlines, linetype = 2, linewidth = 0.6, alpha = 0.8)
  
  p_top <- ggplot(df, aes(time, data)) +
    geom_point(size = point_size, shape = 21, alpha = 0.5) +
    geom_line(aes(y = trend), linewidth = 1, color = "red", alpha = 0.8) +
    vline_layer +
    theme_bw() +
    labs(title = title, x = NULL, y = "trend")
  
  p_tcp <- ggplot(df, aes(time, tcpPr)) +
    geom_ribbon(aes(ymin = 0, ymax = tcpPr), fill = "darkgreen", alpha = 0.2) +
    geom_line(linewidth = 0.25, color = "black") +
    vline_layer +
    theme_bw() +
    labs(x = NULL, y = "Pr(tcp)")
  
  p_order <- ggplot(df, aes(time, orderT)) +
    geom_ribbon(aes(ymin = 0, ymax = orderT), fill = "darkgreen", alpha = 0.2) +
    geom_line(linewidth = 0.25, color = "black") +
    theme_bw() +
    labs(x = NULL, y = "orderT")
  
  p_slp <- ggplot(slp_df, aes(time, prob, fill = sign)) +
    geom_area(position = "stack", color = "black", linewidth = 0.2) +
    theme_bw() +
    labs(x = NULL, y = "slpSign") +
    guides(fill = "none")
  
  p_err <- ggplot(df, aes(time, err)) +
    geom_hline(yintercept = 0, linewidth = 0.4) +
    geom_col(width = 0.2, linewidth = 0.1) +
    theme_bw() +
    labs(x = "Time", y = "error")
  
  list(
    plot = (p_top / p_tcp / p_order / p_slp / p_err),
    cp_tbl = cp_tbl,
    vlines = vlines
  )
}

x_sst  <- ggbeast(beast_sst,  title = "SST")
x_ssta <- ggbeast(beast_ssta, title = "SSTA")
x_dhw  <- ggbeast(beast_dhw,  title = "DHW")

fig_s1 <- (x_sst$plot | x_ssta$plot | x_dhw$plot)
fig_s1


```

### Figure S2 Spatiotemporal patterns of maximum DHW

Annual maximum DHW mapped per year across the GBR domain, 1981–2026, on a common 0–21 °C-week scale. The early record is dominated by low-DHW (blue) years with isolated moderate events, transitioning to frequent high-DHW years from 2016 onward; 2016 and 2017 load heavily on the far northern GBR, 2020 spreads broadly with the strongest northern and central signal, and 2024–2025 show extensive high DHW concentrated in the north. The panel sequence makes the temporal escalation and its northward focus simultaneously legible, showing that recent events are not only hotter but spatially more extensive than any in the preceding three decades, and that the northern GBR bears the most severe and repeated accumulated thermal stress.


```{r, figure_s2}
#| message: false
#| warning: false
#| cache: false
#| fig-height: 10
#| fig-width: 12

gbr_OISST2_dhw_annual_max <- summarise_raster(rast("~/GBR-dhw/datasets/GBR_OISST_DHW.tif"),
                                              index = "years", fun = max, na.rm = TRUE, overwrite=TRUE)


names(gbr_OISST2_dhw_annual_max) <- sub("^y_", "", names(gbr_OISST2_dhw_annual_max))

dhw_numeric_breaks <- c(0, 3, 6, 9, 12, 15, 18, 21)  # Adjust numeric breaks
dhw_colour_breaks <- c("#006f99", "#00A6E5", "#FFD700", "#FF8C00", "#B20000", "#660000", "#3D0000")

Figure_S2a <- ggplot() + 
  theme_bw() +
  
  facet_wrap(~lyr, ncol = 12) +
  tidyterra::geom_spatraster(data = gbr_OISST2_dhw_annual_max) +
  # scale_fill_distiller(
  #   palette = "YlOrRd", 
  #   direction = 1, 
  #   limits = c(0, 25), 
  #   na.value = "azure",
  #   name="DHW"
  # ) +
  scale_fill_gradientn(
    limits = c(0, 21),
    name = expression(DHW[max]),
    na.value = "transparent",
    breaks = dhw_numeric_breaks,
    labels = dhw_numeric_breaks,
    colours = dhw_colour_breaks,
    values = scales::rescale(dhw_numeric_breaks, from = c(0, 20))
  ) +
  tidyterra::geom_spatvector(data = aus_outline, fill = "beige") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "bottom",                       # Place legend below the plot
    legend.box = "horizontal",                        # Arrange legend items horizontally
    legend.key.width = unit(1.5, "cm"),               # Adjust legend key width
  #  legend.margin = margin(t = 5, b = 5)              # Add vertical margins to legend
  )

Figure_S2a


```

### Figure S3 Reef-level annual maximum Degree Heating Weeks by latitude.

A heatmap of per-reef annual maximum DHW (colour, 0–21 °C-weeks) with every reef ordered by latitude on the y-axis (−9° north to −25° south) against year on the x-axis. Vertical warm bands mark the mass-bleaching years (1998, 2002, 2016–2025), which intensify and broaden through time, while the most intense DHW (dark red, >15 °C-weeks) concentrates between roughly −12° and −15° in 2016–2024. The figure shows that thermal stress is organised as coherent basin-scale episodes affecting large contiguous latitudinal bands rather than as fine reef-scale noise, directly supporting the model's regional (~357 km) spatial field, and that the northern-central GBR has experienced both the most frequent and the most extreme recent heat exposure.


```{r fig_s3_heatmap}
#| message: false
#| warning: false
#| cache: true
#| fig-height: 15
#| fig-width: 8

dhw_numeric_breaks <- c(0, 3, 6, 9, 12, 15, 18, 21)  # Adjust numeric breaks
dhw_colour_breaks <- c("#006f99", "#00A6E5", "#FFD700", "#FF8C00", "#B20000", "#660000", "#3D0000")

GBR_OISST_DHW <- rast("~/GBR-dhw/datasets/GBR_OISST_DHW.tif")

# maxSST per year for OISST in rast
gbr_OISST2_dhw_annual_max <- summarise_raster(GBR_OISST_DHW,
                                              index = "years", fun = max, na.rm = TRUE, overwrite=TRUE)




### OISST2

# modified function from DHW:

extract_reefs2 <- function (input, shpfile, output = "sf", fun = "mean", weights = "area", 
    timeseries = "daily", varname = "sst") 
{
    input <- terra::project(input, sf::st_crs(shpfile)$wkt)
    names(input) <- terra::time(input)
    extracted_output <- exactextractr::exact_extract(input, shpfile, 
        progress = FALSE, fun = fun, weights = weights, append_cols = c("LABEL_ID", 
            "GBR_NAME"))
    output_file <- dplyr::mutate(tidyr::pivot_longer(extracted_output, 
        cols = -c("LABEL_ID", "GBR_NAME"), names_to = "date", 
        values_to = varname), date = sub("^mean\\.", "", date))
    if (output == "sf") {
        output_file <- shpfile %>% dplyr::left_join(., output_file, 
            by = c("LABEL_ID", "GBR_NAME"))
    }
    return(output_file)
}

suppressMessages(suppressWarnings(
GBR_OISST2_DHW_annual_reefs <- extract_reefs2(gbr_OISST2_dhw_annual_max, gbr_reefs) |>
  st_centroid() |>
  mutate(lon = st_coordinates(geometry)[,1],
         lat = st_coordinates(geometry)[,2]) |>
  st_drop_geometry() |>
  as.data.frame() |>
  na.omit() |>
  mutate(y = dense_rank(desc(LABEL_ID))) |>
  mutate(year = as.numeric(date)) |>
  select(-date)
))

ggplot() + theme_bw() +
  geom_tile(data=GBR_OISST2_DHW_annual_reefs, aes(x=year, y=y, color=sst)) +
  xlab("Year") + ylab("Latitude") +
  scale_x_continuous(
    expand = c(0, 0),
    limits = c(1980, 2026),
    breaks = seq(1980, 2026, 2)
  ) +
  scale_y_continuous(
    expand = c(0, 0),
    breaks = seq(floor(min(GBR_OISST2_DHW_annual_reefs$y)), ceiling(max(GBR_OISST2_DHW_annual_reefs$y)), length.out = 9),
    labels = seq(floor(min(GBR_OISST2_DHW_annual_reefs$lat)), ceiling(max(GBR_OISST2_DHW_annual_reefs$lat)), length.out = 9)
  ) +
  scale_color_gradientn(
    limits = c(0, 21),
    name = expression(DHW[max]),
    na.value = "transparent",
    breaks = dhw_numeric_breaks,
    labels = dhw_numeric_breaks,
    colours = dhw_colour_breaks,
    values = scales::rescale(dhw_numeric_breaks, from = c(0, 20))
  ) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1)
  ) + ggtitle("")



```

### Figure S4 Thermal profiles across DHW event

Seasonal DHW accumulation trajectories (November–August) for each mass-bleaching event, with the long-term climatological mean (black, grey ribbon) for reference and dashed horizontal lines marking each event's peak DHW; the two panels contrast different reference framings/regions. Events differ markedly in both magnitude and timing — 2020 reaches the highest peak (~9.4 DHW) with a late-summer maximum, while 2024 and 2025 (~7 DHW) plateau earlier — and all recent events sit far above the climatological envelope. The figure demonstrates that "DHW" summarises heterogeneous thermal histories: events of similar peak DHW can differ in onset, duration and shape, which is the basis for the dose–response varying among years and a caution that peak DHW alone does not capture the full thermal exposure driving bleaching.

```{r}
#| message: false
#| warning: false
#| cache: true
#| fig-height: 7
#| fig-width: 12


GBR_OISST_DHW <- rast("~/GBR-dhw/datasets/GBR_OISST_DHW.tif")
GBR_CRW_DHW <- rast("~/GBR-dhw/datasets/GBR_CoralTemp_DHW.tif")



annual_OISST <- dhw::plot_annual_dhw(GBR_OISST_DHW)
annual_CRW <- dhw::plot_annual_dhw(GBR_CRW_DHW)



annual_OISST + scale_y_continuous(limits=c(0,10), breaks=seq(0,10,2)) | 
  annual_CRW + scale_y_continuous(limits=c(0,10), breaks=seq(0,10,2)) 



```

### Figure S5 INLA spatial mesh and survey footprints 

The constrained Delaunay triangulation used for the `SPDE` spatial field, with the GBRMP domain boundary and, per event year, the reef locations actually surveyed (coloured points). The mesh covers the full reef domain at consistent resolution (inner edge 40 km, outer extension to 800 km, `EPSG:32755`), while the survey footprints vary substantially among years — 1998/2002 span the length of the reef, 2016/2017 emphasise the north, and 2025 is restricted to a narrow central band. This figure is the visual justification for the entire repeat-survey and composition-control analysis: because which reefs were surveyed differs systematically by year, pooled year effects risk confounding temporal change with survey footprint, motivating both the spatial field and the within-reef analyses.

```{r inlamesh}
#| message: false
#| warning: false
#| cache: true
#| fig-height: 7
#| fig-width: 10


library(INLA)
library(fmesher)

gbr_reef_outline <- readRDS("~/GBR-dhw/datasets/gbr_reef_outline.rds")

GBRMPpolygon <- readRDS("~/GBR-dhw/data/GBRMPpolygon.rds")


GBRMPpolygon_complex <- gbr_reef_outline |>
  dplyr::filter(FEAT_NAME == "Reef") |>
  concaveman::concaveman(concavity = 2, length_threshold = 0) |>
  st_zm() |> st_transform(32755) |> st_make_valid() |> st_buffer(10000)

GBRMPpolygon <- GBRMPpolygon_complex |> st_simplify(dTolerance = 10000, preserveTopology = FALSE)

pts  <- GBRMPpolygon |> st_coordinates() |> as.data.frame() |> select(1,2) |> as.matrix()

mesh <- inla.mesh.2d(loc.domain = pts[,1:2], boundary = GBRMPpolygon, max.edge = c(40000, 800000), offset = c(5000, 10000), crs="EPSG:32755")


dhw_inla <- readRDS("~/GBR-dhw/datasets/gbr_shape_aerial_OISST.rds") |>
  st_transform(32755) |>
  mutate(aerial = case_when(
    max_aerial_score == 0 ~ "0",
    max_aerial_score %in% c(1, 2) ~ "1",
    max_aerial_score == 3 ~ "2",
    max_aerial_score %in% c(4,5) ~ "3",
    TRUE ~ NA_character_  
  ) ) |>
  mutate(aerial = as.numeric(as.character(aerial)) + 1) |>  # note - +1 val
  select(-max_aerial_score, -median_aerial_score)


ggplot() + theme_bw() +
  facet_wrap(~year, nrow=2) +
  ggtitle("GBRMP mesh for aerial surveys by year") +
  geom_sf(data = fm_as_sfc(mesh, multi = FALSE) |> st_make_valid() |> st_union(), linewidth=0.25, fill=NA) +
  geom_sf(data = fm_as_sfc(mesh, multi = FALSE), linewidth=0.04, fill=NA, color="grey") +
  geom_sf(data=dhw_inla, aes(fill=as.factor(year), color=as.factor(year)), alpha=0.2, stroke=1, size = 2, show.legend=FALSE) +
  scale_fill_brewer(palette="RdBu", direction=-1) +
  scale_color_brewer(palette="RdBu", direction=-1) +
  geom_sf(data = GBRMPpolygon, linewidth=0.1, fill=NA, color="blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position="bottom")

```

### Figure S6 Latent-scale predictive uncertainty

Maps of the 95% credible-interval width of the latent linear predictor η (logit scale) across the hexagonal prediction grid for each event. Uncertainty is lowest (pale) along the densely surveyed reef matrix and rises (dark) at the domain margins and sparsely sampled extremities, with the spatial pattern essentially constant across years. Reporting uncertainty on the latent scale rather than the probability scale avoids the artificial compression the logistic transformation imposes near p = 0.5, and the figure shows that predictions are well constrained where survey density is high and should be interpreted cautiously at the poorly sampled edges — consistent with the decision to interpret predicted surfaces at the regional rather than the grid-cell scale.
                        
```{r}
#| message: false
#| warning: false
#| cache: true
#| fig-height: 5
#| fig-width: 12

GBR_grid_pred <- readRDS("~/GBR-dhw/data/GBR_grid_pred.rds")

p_eta_unc <- ggplot(GBR_grid_pred) +
  geom_sf(aes(fill = eta_ci_width), colour = NA) +
  scale_fill_viridis_c(option = "mako", direction = -1,
                       name = "Latent 95%\nCI width (logit)") +
  facet_wrap(~ year, nrow = 2) +

  theme_void() +
  theme(legend.position = "right", strip.text = element_text(size = 9),
        plot.subtitle = element_text(size = 8))
 
p_eta_unc
 


```

### Figure S7 Modal ordinal predictions

The most-likely bleaching category per grid cell per event, mapped on the collapsed ordinal scale (No/low, Low–moderate, Moderate, High/severe. The modal surface shifts from predominantly No/low in 1998/2002 to extensive High/severe and Moderate classes in 2016, 2020 and 2022, with large NA areas in years of restricted survey coverage. The figure translates the latent model into an interpretable categorical map, showing the regional progression toward more severe modal states in recent events while honestly masking cells where the prediction would be an extrapolation beyond sampled DHW.

```{r}
#| message: false
#| warning: false
#| cache: true
#| fig-height: 4
#| fig-width: 12

#  - cat_mode = the model's single best-guess bleaching class per cell -> a
#    categorical analogue of the raw interpolated Fig 2a, but model-driven.

GBRMPpolygon <- readRDS("~/GBR-dhw/data/GBRMPpolygon.rds")
GBR_grid_pred_pom <- readRDS("~/GBR-dhw/data/GBR_grid_pred_pom.rds")


# ---- discrete (categorical) map ---------------------------------------------
cat_vals <- c("1"="#4DADE6","2"="#5BC36E","3"="#FFFF80","4"="#FF784E")
cat_labs <- c("1"="No/low","2"="Low–moderate","3"="Moderate","4"="High/severe")

p_main <- ggplot() +
  geom_sf(data = GBRMPpolygon, linewidth = 0.25, colour = "lightgrey", fill = NA) +
  geom_sf(data = GBR_grid_pred_pom, aes(fill = cat_mode), colour = NA) +
  geom_sf(data = GBRMPpolygon, linewidth = 0.25, colour = "black", fill = NA) +
  scale_fill_manual(values = cat_vals, labels = cat_labs, name = "Most likely\ncategory",
                    drop = FALSE) +
  facet_wrap(~ year, nrow = 2) +
  theme_void() + theme(legend.position = "right", strip.text = element_text(size = 9))

p_main

```
### Figure S8 Probability of severe bleaching for INLA spatial model

Mapped posterior-mean `P(Y = severe) = 1 − F(α₃ − η)` per grid cell per event, on a 0–100% scale. Severe-bleaching probability is low and spatially patchy in 1998/2002, becomes high and widespread from 2016 onward, and reaches its most extensive coverage in 2020, 2022 and 2024, with the northern and central reef matrix most affected. The figure is the probability-scale expression of the headline result — the regional likelihood of severe bleaching at observed thermal stress has risen markedly across events — and complements the latent-scale uncertainty map (S6) that qualifies where these probabilities are well constrained.


```{r}
#| message: false
#| warning: false
#| cache: true
#| fig-height: 4
#| fig-width: 12

library(scales)

sev_pal <- c("#2c7bb6","#ffffbf","#fdae61","#d7191c","darkred")  # blue->red

GBR_grid_pred <- readRDS("~/GBR-dhw/data/GBR_grid_pred.rds")
GBRMPpolygon <- readRDS("~/GBR-dhw/data/GBRMPpolygon_grid_pred.rds")

p_main <- ggplot() +
  geom_sf(data = GBRMPpolygon, linewidth = 0.25, color = "lightgrey", fill = NA) +
  geom_sf(data = GBR_grid_pred, aes(fill = p_severe), show.legend=TRUE, colour = NA) +
 # geom_sf(data = GBRMPpolygon2025, linewidth = 0.25, color = "black", fill = NA) +
  geom_sf(data = GBRMPpolygon, linewidth = 0.25, color = "black", fill = NA) +
  scale_fill_gradientn(colours = sev_pal, limits = c(0,1),
                       name = "P(severe)", labels = percent) +
  facet_wrap(~ year, nrow = 2) +
  # labs(title = "Model-predicted probability of severe bleaching",
  #      subtitle = "Posterior mean from the spatial POM model (SPDE field + DHW + year)") +
  theme_void() + theme(legend.position = "right",
                       strip.text = element_text(size = 9))


p_main


```

### Figure S9 DHW predictions (NOAA CRW)

Panel A tracks the 30 reefs surveyed in all eight events (1998–2025) through their ordered bleaching category, with ribbons tracing individual-reef transitions and per-event maps of scored reefs; panel B shows the larger 2016–2025 panel (n = 81). The alluvial flows show increasing occupancy of high categories (4–5) in recent events and a rising frequency of upward transitions, while the 2022 column carries the only substantial category-5 (most severe) block. The figure presents the repeat-survey evidence on the full (non-balanced) dataset, demonstrating that the erosion of recovery and accumulation in severe categories is visible across both the long (1998-anchored, n = 30) and recent (2016-anchored, n = 81) repeat panels, and it is the data source underlying the within-reef contextual-vulnerability analyses.

```{r}
#| message: false
#| warning: false
#| cache: true
#| fig-height: 4
#| fig-width: 12

library(sf)
library(tidyverse)
library(INLA)
library(concaveman)
library(fmesher)
library(terra)
library(kableExtra)
library(patchwork)
library(ggplot2)
library(ggridges)



## load observation data

dhw_inla_crw <- readRDS("~/GBR-dhw/datasets/gbr_shape_aerial_CRW.rds") |>
  st_transform(32755) |>
  mutate(aerial = case_when(
    max_aerial_score == 0 ~ "0",
    max_aerial_score == 1 ~ "1",
    max_aerial_score == 2 ~ "2",
    max_aerial_score == 3 ~ "3",
    max_aerial_score == 4 ~ "4",
    max_aerial_score == 5~ "4",
    TRUE ~ NA_character_
  ) ) |>
  mutate(aerial = as.numeric(as.character(aerial)) + 1) |>  # note - +1 val
  select(-max_aerial_score, -median_aerial_score)

# load boundary
# GBRMPpolygon_complex <- st_read("~/GBR-dhw/archive/datasets/aerialshp/TS_GBRMP_FEATURES.shp", quiet=TRUE) |>

gbr_reef_outline <- readRDS("~/GBR-dhw/datasets/gbr_reef_outline.rds")


GBRMPpolygon_complex <- gbr_reef_outline |>
  dplyr::filter(FEAT_NAME=="Reef") |>
  concaveman::concaveman(concavity = 2, length_threshold = 0) |>
  st_zm() |>
  st_transform(32755) |>
  st_make_valid() |>
  st_buffer(10000)

GBRMPpolygon <- GBRMPpolygon_complex |> st_simplify(dTolerance = 10000, preserveTopology = FALSE)

# make prediction grid
GBRMPpredict_grid_CRW <- st_make_grid(st_bbox(GBRMPpolygon), cellsize = 27830, square = FALSE) |>
  st_as_sf() |>
  st_filter(GBRMPpolygon, .predicate = st_intersects)

# CRW_dhw <- rast("~/GBR-dhw/datasets/GBR_CoralTemp_DHW.tif")
# CRW_dhw <- project(CRW_dhw, "EPSG:32755")
# CRW_annual_max_dhw <- terra::tapp(CRW_dhw, index = "years", fun = max, na.rm = TRUE, overwrite=TRUE)
# CRW_annual_max_dhw <- CRW_annual_max_dhw[[paste0("y_",  c(1998, 2002, 2016, 2017, 2020, 2022, 2024, 2025))]]
# names(CRW_annual_max_dhw) <- sub("^y_", "", names(CRW_annual_max_dhw))
# writeRaster(CRW_annual_max_dhw, "~/GBR-dhw/datasets/CRW_annual_max_dhw.tif")
CRW_annual_max_dhw <- rast("~/GBR-dhw/datasets/CRW_annual_max_dhw.tif")

names(CRW_annual_max_dhw) <- c(1998, 2002, 2016, 2017, 2020, 2022, 2024, 2025)

# update with processed CRW
GBRMP_aerial_CRW <- dhw_inla_crw  |> arrange(year, aerial) |> mutate(year=factor(year, levels = c(1998, 2002, 2016, 2017, 2020, 2022, 2024, 2025)))
GBRMP_aerial_CRW_centroids <- GBRMP_aerial_CRW |> st_centroid()
GBRMP_aerial_CRW_centroids$dhw <- GBRMP_aerial_CRW_centroids$dhwCRW

# extract DHW for the prediction grid and convert to points
GBRMPpredict_grid_dhw_CRW <- list()
for (i in 1:8) {
  #for (i in 1:length(names(CRW_annual_max_dhw))) {
  GBRMPpredict_grid_CRW$dhw <- exactextractr::exact_extract(CRW_annual_max_dhw[[i]], GBRMPpredict_grid_CRW, fun = "max", progress = FALSE, weights = "area")
  GBRMPpredict_grid_dhw_CRW[[i]] <- GBRMPpredict_grid_CRW |> mutate(year = as.factor(names(CRW_annual_max_dhw)[[i]]))
}
GBRMPpredict_grid_dhw_CRW <- do.call(rbind, GBRMPpredict_grid_dhw_CRW)
GBR_grid_CRW <- GBRMPpredict_grid_dhw_CRW |> na.omit()
GBR_grid_CRW_centroids <- GBRMPpredict_grid_dhw_CRW |> st_centroid() |> na.omit()

# make points along linestring for border extent
boundary_CRW <- GBR_grid_CRW_centroids |>
  select(-year, -dhw) |>
  distinct() |>
  concaveman::concaveman() |>
  st_buffer(10000) |>
  st_boundary()

# get boundary points and make mesh
pts_CRW <- GBRMPpolygon |> st_coordinates() |> as.data.frame() |> select(1,2) |> as.matrix()

# max.edge is the largest allowed triangle length; the lower the number the higher the resolution max.edge = c(inside the boundary triangle, outside the boundary triangle).
# offset is defining how far you want to extend your domain (i.e. a secondary boundary box) for inner boundary and outer boundary; the offset goes with the max edge
# see https://sites.stat.washington.edu/peter/591/Lindgren.pdf

mesh_CRW <- fmesher::fm_mesh_2d_inla(loc.domain = pts_CRW[,1:2], boundary = GBRMPpolygon, max.edge = c(40000, 800000), offset = c(5000, 10000), crs="EPSG:32755")

# Define the SPDE model using the mesh
#spde_CRW <- inla.spde2.matern(mesh = mesh_CRW, alpha = 2)

spde_CRW <- inla.spde2.pcmatern(
  mesh_CRW,
  alpha = 2,
  prior.range = c(150000, 0.5),   # P(range < 150 km) = 0.5   -- metres (EPSG:32755)
  prior.sigma = c(3, 0.01))       # P(sigma > 3) = 0.01  -- field SD ~3.7 was observed, so allow it

# Create a projection matrix for the observed data
A_obs_CRW <- inla.spde.make.A(mesh = mesh_CRW, loc = st_coordinates(GBRMP_aerial_CRW_centroids))

# Create an index for the spatial field (random effect) in the SPDE model
s_index_CRW <- inla.spde.make.index(name = "spatial.field", n.spde = spde_CRW$n.spde)

# Prepare the data stack for the observed  data
obs_stack_CRW <- inla.stack(
  data = list(aerial = GBRMP_aerial_CRW_centroids$aerial),  # Response variable
  A = list(A_obs_CRW, 1),  # Projection matrix for the spatial model and fixed effects
  effects = list(
    c(s_index_CRW, list(Intercept = 1)),  # Spatial random effects + intercept
    list(dhw = GBRMP_aerial_CRW_centroids$dhw, year = GBRMP_aerial_CRW_centroids$year, LABEL_ID = GBRMP_aerial_CRW_centroids$LABEL_ID)  # Covariates: dhw and year
  ),
  tag = "obs"  # Tag for the observed data stack
)


# repeat expand_grid for each grid-cell layout
min_dhw_CRW <- 0  # Minimum DHW
max_dhw_CRW <- 16
dhw_seq_CRW <- seq(min_dhw_CRW, max_dhw_CRW, length.out = 17)  # Generate 100 values between min and max DHW
unique_years_CRW <- unique(GBR_grid_CRW_centroids$year)
pred_grid_CRW <- expand.grid(dhw = dhw_seq_CRW, year = unique_years_CRW) # Create a grid of dhw and year combinations
n_locs_CRW <- nrow(GBR_grid_CRW_centroids) # Number of spatial locations
expanded_coords_CRW <- st_coordinates(GBR_grid_CRW_centroids)[rep(1:n_locs_CRW, times = nrow(pred_grid_CRW)), ] # Replicate the spatial locations for each dhw-year combination

# make preds mesh / stack
A_pred_expanded_CRW <- inla.spde.make.A(mesh = mesh_CRW, loc = expanded_coords_CRW)
expanded_grid_CRW <- pred_grid_CRW[rep(1:nrow(pred_grid_CRW), each = n_locs_CRW), ]
expanded_grid_sf_CRW <- cbind(expanded_grid_CRW, expanded_coords_CRW) |> st_as_sf(coords=c("X", "Y"), crs=32755)

preds_stack_CRW <- inla.stack(
  data = list(aerial = NA),  # NA for predictions
  A = list(A_pred_expanded_CRW, 1),
  effects = list(
    c(s_index_CRW, list(Intercept = 1)),  # Spatial random effects + intercept for prediction
    list(dhw = expanded_grid_CRW$dhw, year = expanded_grid_CRW$year, LABEL_ID = NA)  # Expanded dhw and year
  ),
  tag = "preds"
)

# polygon prediction stack


GBRMP_aerial_centroids <- GBRMP_aerial_CRW |> st_centroid() |> mutate(dhw = dhwCRW)

gbr_reef_subset <- gbr_reef_outline |>
  st_zm() |>
  filter(!FEAT_NAME %in% c("Mainland", "Other")) |>
  filter(X_COORD > 140) |>
  filter(Y_COORD > -24.3) |>
  st_make_valid() |>
  st_transform(4326)


gbr_reef_32755_CRW <- gbr_reef_subset |>          # <- the reef POLYGON layer
  st_make_valid() |> st_transform(32755)

yrs_CRW <- names(CRW_annual_max_dhw)            # "1998","2002",...,"2025"

reef_year_dhw_CRW <- lapply(seq_along(yrs_CRW), function(i) {
  dhw_i <- exactextractr::exact_extract(
    CRW_annual_max_dhw[[i]],               # year i's DHW raster layer
    gbr_reef_32755_CRW,                    # extract onto each reef polygon
    fun = "max", progress = FALSE)
  gbr_reef_32755_CRW |>
    transmute(TARGET_FID, LABEL_ID, year = yrs_CRW[i], dhw = dhw_i)
}) |>
  bind_rows() |>
  filter(is.finite(dhw))

pred_pts_CRW <- reef_year_dhw_CRW |>
  st_centroid()

year_levels_CRW <- union(levels(GBRMP_aerial_CRW_centroids$year), sort(unique(pred_pts_CRW$year)))
id_levels_CRW   <- union(levels(factor(GBRMP_aerial_CRW_centroids$LABEL_ID)), sort(unique(pred_pts_CRW$LABEL_ID)))

pred_pts2_CRW <- pred_pts_CRW |>
  mutate(
    year = factor(as.character(year), levels = year_levels_CRW),
    LABEL_ID = factor(as.character(LABEL_ID), levels = id_levels_CRW)
  )

# build reef_year_dhw_CRW + pred_pts2_CRW with harmonised factor levels as before, then:
pred_stack_poly_CRW <- inla.stack(
  data = list(aerial = NA),
  A = list(inla.spde.make.A(mesh_CRW, st_coordinates(pred_pts2_CRW)), 1),
  effects = list(c(s_index_CRW, list(Intercept = 1)),
                 list(dhw = pred_pts2_CRW$dhw, year = pred_pts2_CRW$year, 
                      LABEL_ID = pred_pts2_CRW$LABEL_ID)),
  tag = "pred_poly")

combined_stack_CRW <- inla.stack(obs_stack_CRW, preds_stack_CRW, pred_stack_poly_CRW)   


# Define the INLA formula (full spatial model)
inla_formula_CRW <- aerial ~ dhw * year + f(spatial.field, model = spde_CRW) + f(LABEL_ID, model = "iid")

# Define fixed effect priors
fixedpriors_CRW <- list(dic = TRUE, waic = TRUE, config = TRUE, return.marginals.predictor = TRUE)

fm1_CRW <- inla(
  inla_formula_CRW,
  data = inla.stack.data(combined_stack_CRW),
  control.predictor = list(A = inla.stack.A(combined_stack_CRW), 
                           compute = TRUE, link = 1),
  family = "pom",
  control.family = list(control.pom = list(cdf = "logit")),
  control.fixed = list(
    mean.intercept = -2,        # Baseline very low bleaching
    prec.intercept = 1,         # Moderate certainty in that belief
    mean = c(dhw = 1, year = 0),
    prec = c(dhw = 0.001, year = 0.001)  # weak dhw prior, weak year prior
  ),
  control.compute =  list(dic = TRUE, waic = TRUE, 
                          config = TRUE, return.marginals.predictor = TRUE),
  control.inla = list(control.vb = list(emergency = 30)),
  silent=2L
)

# saveRDS(fm1_CRW, "~/GBR-dhw/data/fm1_CRW.rds")
# fm1_CRW <- readRDS("~/GBR-dhw/data/fm1_CRW.rds")

# sp_CRW <- inla.spde2.result(fm1_CRW, "spatial.field", spde_CRW, do.transform = TRUE)
# inla.qmarginal(c(.025,.5,.975), sp_CRW$marginals.range.nominal[[1]])   # range, metres
# inla.qmarginal(c(.025,.5,.975), sp_CRW$marginals.variance.nominal[[1]]) # marginal variance


##### extract fixed effects

# Extract marginals from fm1_CRW$marginals.fixed and convert them into a data frame
combined_df_fixef_CRW <- bind_rows(
  lapply(names(fm1_CRW$marginals.fixed), function(param) {
    df <- as.data.frame(fm1_CRW$marginals.fixed[[param]])
    df$parameter <- param
    return(df)
  })
) |>
  rename(x=1, height=2, parameter=3)

# Extract marginals from fm1_CRW$marginals.fixed and convert them into a data frame
combined_df_hyper_CRW <- bind_rows(
  lapply(names(fm1_CRW$marginals.hyperpar), function(param) {
    df <- as.data.frame(fm1_CRW$marginals.hyperpar[[param]]) # Convert to data frame
    df$parameter <- param # Add a column for the parameter name
    return(df)
  })
) |>
  rename(x=1, height=2, parameter=3) |>
  mutate(parameter = factor(parameter,
                            levels=c("theta1 for POM",
                                     "theta2 for POM",
                                     "theta3 for POM",
                                     "theta4 for POM",
                                     "Theta1 for spatial.field",
                                     "Theta2 for spatial.field")))



# Extract marginals from fm1_CRW$marginals.fixed and convert them into a data frame
combined_df_fixef_CRW <- bind_rows(
  lapply(names(fm1_CRW$marginals.fixed), function(param) {
    df <- as.data.frame(fm1_CRW$marginals.fixed[[param]])
    df$parameter <- param
    return(df)
  })
) |>
  rename(x = 1, height = 2, parameter = 3) |>
  filter(!parameter == "(Intercept)") |>
  mutate(parameter = factor(parameter, levels = c(
    "dhw", "year2002", "year2016", "year2017", "year2020", "year2022", "year2024", "year2025",
    "dhw:year2002", "dhw:year2016", "dhw:year2017", "dhw:year2020", "dhw:year2022",
    "dhw:year2024","dhw:year2025"
  ), ordered = TRUE))



fixef_plot_ms_CRW <- ggplot() + 
  geom_ridgeline(data=combined_df_fixef_CRW, 
                 aes(x = x, y = fct_rev(parameter), height = height/10, group = parameter), 
                 fill = "aquamarine4", alpha=0.5) +
  theme_bw() +
  xlim(-3,6) + ylab("") +
  geom_vline(xintercept=0, linewidth=0.5, color="darkgrey") +
  #labs(title = "Posterior Marginals for Fixed Effects", x = "Value (x)", y = "Parameter") +
  theme(legend.position = "none") +
  xlab("Posterior Mean (Effect Size)")


# ordinal (POM) prediction + stacked cumulative-probability figure
# model was fit with control.pom = list(cdf = "logit"), so the back-transform 
# must use plogis(). cut-points are reconstructed from INLA's internal increment
# parameterisation (alpha1 = theta1; alpha_k = alpha_{k-1} + exp(theta_k)),
# i.e.correct ordered-threshold scale 


# ---- 1. POM category probabilities from latent eta + ordered cut-points -----
# P(Y = k) = F(alpha_k - eta) - F(alpha_{k-1} - eta), with F = plogis (logit)
predict_pom_with_bounds_CRW <- function(eta_mean, eta_lower, eta_upper, cutpoints,
                                        link = "logit") {
  cdf <- switch(link, logit = plogis, probit = pnorm,
                stop("link must be 'logit' or 'probit'"))
  cutpoints <- sort(cutpoints)                 # ensure ordered thresholds
  n <- length(eta_mean)
  K <- length(cutpoints) + 1L
  
  cat_probs <- function(eta) {
    p <- diff(c(0, cdf(cutpoints - eta), 1))   # K categories
    pmax(p, 0)
  }
  P  <- t(vapply(eta_mean,  cat_probs, numeric(K)))
  Pl <- t(vapply(eta_lower, cat_probs, numeric(K)))
  Pu <- t(vapply(eta_upper, cat_probs, numeric(K)))
  
  list(mean  = as.data.frame(P),
       lower = as.data.frame(pmin(Pl, Pu)),    # CI bounds can invert per category
       upper = as.data.frame(pmax(Pl, Pu)))
}

# ---- 2. reconstruct ORDERED cut-points (not raw thetas) ---------------------
th_CRW <- c(fm1_CRW$summary.hyperpar["theta1 for POM", "mean"],
            fm1_CRW$summary.hyperpar["theta2 for POM", "mean"],
            fm1_CRW$summary.hyperpar["theta3 for POM", "mean"],
            fm1_CRW$summary.hyperpar["theta4 for POM", "mean"])
cutpoints_CRW <- cumsum(c(th_CRW[1], exp(th_CRW[-1])))     # alpha1, alpha1+exp(th2), ...
# (these should be increasing, e.g. ~ -2.57, 0.46, 1.82)

# ---- 3. use the LATENT LINEAR PREDICTOR, not fitted values ------------------
lp_CRW <- fm1_CRW$summary.linear.predictor            # <- eta scale, what POM needs
eta_mean_CRW  <- lp_CRW$mean
eta_lower_CRW <- lp_CRW$`0.025quant`
eta_upper_CRW <- lp_CRW$`0.975quant`

pp_CRW <- predict_pom_with_bounds_CRW(eta_mean_CRW, eta_lower_CRW, eta_upper_CRW, cutpoints_CRW,
                                      link = "logit")

# ---- 4. attach to the prediction grid ---------------------------------------
index_pred_CRW <- inla.stack.index(combined_stack_CRW, tag = "preds")$data
point_data_CRW <- expanded_grid_CRW |>
  mutate(loc_id = rep(seq_len(n_locs_CRW), times = nrow(pred_grid_CRW)))

predicted_ordinal_mean_CRW <- cbind(point_data_CRW, pp_CRW$mean[index_pred_CRW, ]) |>
  pivot_longer(cols = starts_with("V"), names_to = "aerial_score", values_to = "mean") |>
  mutate(
    aerial = recode(
      aerial_score,
      V1 = "No bleaching",
      V2 = "Low bleaching",
      V3 = "Moderate bleaching",
      V4 = "High bleaching",
      V5 = "Very high bleaching"
    ),
    aerial = factor(
      aerial,
      c(
        "No bleaching",
        "Low bleaching",
        "Moderate bleaching",
        "High bleaching",
        "Very high bleaching"
      )
    ),
    year = as.integer(as.character(year))
  )

# ---- 5. mask each panel beyond its observed DHW (no extrapolation shown) -----

max_dhw_by_year_CRW <- GBRMP_aerial_CRW_centroids |> st_drop_geometry() |>
  mutate(year = as.integer(as.character(year))) |>
  group_by(year) |> summarise(min_dhw = ceiling(min(dhw, na.rm = TRUE)), 
                              max_dhw = ceiling(max(dhw, na.rm = TRUE)), .groups = "drop")

# interpcolors <- c("No bleaching"="#74ADD1","Low-Moderate bleaching"="#FEE090",
#                   "High bleaching"="#FDAE61","Very High-Extreme bleaching"="#D73027")


# <- c("No bleaching"="#ABD9E9","Low-Moderate bleaching"="#FFFFBF",
#                  "High bleaching"="#FDAE61","Very High-Extreme bleaching"="#D73027")

interpcolors_aerial_score_full_CRW <- c("#ABD9E9", "#FFFFBF", "#FEE090", "#FDAE61", "#F46D43", "#D73027")
interpcolors_aerial_score_five_CRW <- c("#ABD9E9", "#FFFFBF", "#FDAE61", "#F46D43", "#D73027")



plot_with_mask_CRW <- function(yr) {
  min_dhw <- max_dhw_by_year_CRW |> filter(year == yr) |> pull(min_dhw)
  max_dhw <- max_dhw_by_year_CRW |> filter(year == yr) |> pull(max_dhw)
  predicted_ordinal_mean_CRW_plot <- predicted_ordinal_mean_CRW |>
    group_by(dhw, year, aerial_score, aerial) |>
    summarise(
      mean = mean(mean, na.rm = TRUE),
      .groups = "drop"
    )
  
  ggplot() + theme_bw() +
    geom_area(data = predicted_ordinal_mean_CRW_plot |> filter(year == yr),
              aes(x = dhw, y = mean, fill = aerial),
              position = "stack", colour = "black", linewidth = 0.4,
              show.legend = (yr == 2024)) +
    geom_vline(xintercept = c(2,4,6,8,10,12,14),
               linetype = c("dotted","dotted","dotted","dotdash","dotdash","longdash","longdash"),
               linewidth = 0.2) +
    geom_rect(aes(xmin = 0, xmax = min_dhw, ymin = 0, ymax = 1), fill = "white", alpha=1) +
    geom_rect(aes(xmin = max_dhw, xmax = 16, ymin = 0, ymax = 1), fill = "white") +
    scale_x_continuous(limits = c(0,16), breaks = seq(0,16,4)) +
    scale_fill_manual(values = interpcolors_aerial_score_five_CRW) +
    xlab("") + ylab(if (yr == 1998) "Cumulative probability" else "") +
    ggtitle(paste0(yr, " event")) +
    theme(plot.title = element_text(size = 10),
          plot.margin = unit(c(0,0,0,0), "cm"),
          axis.text.y = if (yr == 1998) element_text() else element_blank(),
          legend.position = if (yr == 2024) "bottom" else "none")
}


years_CRW <- c(1998,2002,2016,2017,2020,2022,2024, 2025)   # note 2025: DHW floor ~7.5, no low-DHW support
plot_list_CRW <- lapply(years_CRW, plot_with_mask_CRW)

pom_preds_CRW <- wrap_plots(plot_list_CRW, ncol = 8) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom", legend.justification = "center")

pom_preds_CRW



```

### Figure S10 DHW comparisons (OISST vs CRW)

The scatter panel (Image 3) cross-plots reef-level CRW DHW against OISST DHW per event with the 1:1 line and per-panel R²; the map panel (Image 6) shows the spatial difference (OISST − CRW) per event. Agreement between products is event-dependent — strong in 2016 (R² = 0.76) and 2022 (R² = 0.60) but weak in 2002 (R² = 0.06) and 2020 (R² = 0.26) — and OISST generally exceeds CRW (points below the 1:1 line; predominantly positive ΔDHW, strongest in the far north). The figure is a product-sensitivity check showing that absolute DHW values are sensitive to the SST product, particularly in the north and in early/low-signal years, which justifies reporting the CRW-based re-analysis (S9) and treating cross-product agreement as a bound on how precisely event thermal dose can be assigned.

```{r}
#| message: false
#| warning: false
#| cache: true
#| fig-height: 6
#| fig-width: 12

library(tidyverse)
library(broom)

gbr_shape_aerial_OISST_comparison <- readRDS("/Users/rof011/GBR-dhw/datasets/gbr_shape_aerial_OISST.rds") |>
  mutate(aerial = max_aerial_score) |>  
  select(-max_aerial_score, -median_aerial_score) |> 
  rename(dhwOISST = dhw)

gbr_shape_aerial_CRW_comparison <- readRDS("~/GBR-dhw/datasets/gbr_shape_aerial_CRW.rds") |>
  mutate(aerial = max_aerial_score) |>  # note - +1 val
  select(-max_aerial_score, -median_aerial_score) 

GBRMP_dhwcomp <- left_join(
    gbr_shape_aerial_OISST_comparison |> as.data.frame() |> select(LABEL_ID, year, dhwOISST),
    gbr_shape_aerial_CRW_comparison |> as.data.frame() |> select(LABEL_ID, year, dhwCRW)
  ) |>
  mutate(
    year = as.integer(as.character(year))
  ) |>
  filter(
    is.finite(dhwOISST),
    is.finite(dhwCRW)
  )


lm_r2_lab <- GBRMP_dhwcomp |>
  group_by(year) |>
  nest() |>
  mutate(
    fit = map(data, ~ lm(dhwCRW ~ dhwOISST, data = .x)),
    glance = map(fit, broom::glance)
  ) |>
  select(year, glance) |>
  unnest(glance) |>
  mutate(
    lab = paste0("R² = ", round(r.squared, 2)),
    x = -Inf,
    y = Inf
  )

dhw_lm_plot <- ggplot() +
  theme_bw() +
  geom_point(
    data = GBRMP_dhwcomp,
    aes(dhwOISST, dhwCRW, fill = factor(year)),
    shape = 21,
    size = 2,
    alpha = 0.5
  ) +
  geom_smooth(
    data = GBRMP_dhwcomp,
    aes(dhwOISST, dhwCRW),
    method = "lm",
    se = TRUE,
    colour = "black",
    fill = "grey70",
    alpha = 0.35,
    linewidth = 0.6
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    colour = "grey40",
    linewidth = 0.5
  ) +
  geom_text(
    data = lm_r2_lab,
    aes(x = x, y = y, label = lab),
    inherit.aes = FALSE,
    hjust = -0.1,
    vjust = 1.3,
    size = 4.5
  ) +
  facet_wrap(~year, nrow=2) +
  labs(
    x = "OISST DHW",
    y = "CRW DHW",
    fill = "Year"
  ) +
  theme(
    legend.position = "none"
  ) + coord_fixed() +
  scale_x_continuous(limits=c(0,20), breaks=seq(0,20,4)) +
  scale_y_continuous(limits=c(0,20), breaks=seq(0,20,4))

dhw_lm_plot


```

### Figure S11 Spatial DHWdelta (OISST - CRW)

Event-specific cumulative-probability (stacked-area) curves of bleaching category against DHW (0–16 °C-weeks) for the proportional-odds model, masked to each event's observed DHW range; colours run No bleaching (blue) through Very high (dark red). The curves shift progressively leftward and upward through time — 1998/2002 retain substantial No/Low probability up to moderate DHW, whereas 2016, 2020, 2022 and 2024 reach high/very-high categories at low DHW — visualising the rising bleaching propensity per unit thermal stress that is the paper's central result.

```{r}
#| message: false
#| warning: false

GBR_OISST_DHW <- rast("~/GBR-dhw/datasets/GBR_OISST_DHW.tif")
GBR_CRW_DHW <- rast("~/GBR-dhw/datasets/GBR_CoralTemp_DHW.tif")

library(tidyverse)
library(terra)
library(tidyterra)
library(ggplot2)

bleaching_years <- c(1998, 2002, 2016, 2017, 2020, 2022, 2024, 2025)

# helper: annual maximum raster for selected years
calc_year_max <- function(x, years) {
  
  x_dates <- as.Date(time(x))
  x_years <- lubridate::year(x_dates)
  
  out <- map(
    years,
    \(yr) {
      
      lyr <- which(x_years == yr)
      
      if (length(lyr) == 0) {
        r <- x[[1]]
        values(r) <- NA_real_
        names(r) <- paste0("y", yr)
        return(r)
      }
      
      r <- app(x[[lyr]], max, na.rm = TRUE)
      names(r) <- paste0("y", yr)
      r
    }
  ) |>
    rast()
  
  time(out) <- as.Date(paste0(years, "-01-01"))
  names(out) <- paste0("y", years)
  
  out
}

# annual maximum DHW
GBR_OISST_DHW_max <- calc_year_max(
  GBR_OISST_DHW,
  bleaching_years
)

GBR_CRW_DHW_max <- calc_year_max(
  GBR_CRW_DHW,
  bleaching_years
)

# project/resample CRW annual maxima onto OISST grid
GBR_CRW_DHW_max_OISSTgrid <- project(
  GBR_CRW_DHW_max,
  GBR_OISST_DHW_max,
  method = "bilinear"
)

names(GBR_CRW_DHW_max_OISSTgrid) <- names(GBR_OISST_DHW_max)
time(GBR_CRW_DHW_max_OISSTgrid) <- time(GBR_OISST_DHW_max)

# delta DHW: positive = OISST higher than CRW
GBR_delta_DHW <- GBR_OISST_DHW_max - GBR_CRW_DHW_max_OISSTgrid

names(GBR_delta_DHW) <- paste0("deltaDHW_", bleaching_years)
time(GBR_delta_DHW) <- as.Date(paste0(bleaching_years, "-01-01"))

deltaDHW_plot <- ggplot() +
  theme_bw() +
  geom_spatraster(
    data = GBR_delta_DHW
  ) +
  facet_wrap(~lyr, ncol = 4) +
  scale_fill_gradient2(
    name = "ΔDHW\nOISST - CRW",
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    na.value = NA
  ) +
  coord_sf(
    xlim = c(141, 152.75),
    ylim = c(-24.5, -9),
    expand = FALSE
  ) +
  labs(
    x = NULL,
    y = NULL
  ) +
  theme(
    strip.background = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right"
  )



deltaDHW_plot


```

### Figure S12 Repeat aerial surveys

Alluvial diagrams track individual reefs through their bleaching category at successive events for reefs surveyed across (A) all eight events (n = 30; 1998–2025) and (B) the 2016–2025 events (n = 81). Bars show the number of reefs in each ordered category per event (0 = none, 5 = most severe); ribbons trace individual-reef transitions. In both panels occupancy of severe categories (4–5) increases through recent events, category 5 appears only from 2022, and the frequency of upward transitions rises across intervals, indicating eroding recovery between events. Per-event maps show the latitudinal distribution of scored reefs.


```{r}
#| message: false
#| warning: false
#| cache: true
#| fig-height: 12
#| fig-width: 12

library(ggalluvial)

path <- "~/GBR-dhw/datasets/gbr_shape_aerial_OISST.rds"
ausmap <- readRDS("~/GBR-dhw/data/ausmap.rds")


plot_alluvial <- function(path,
                           year_min = -Inf,
                           year_max = Inf) {
 

  interpcolors <- c("#4DADE6", "#5BC36E", "#FFFF80",
                    "#FF784E", "#D53B50", "#5A0E0E")

  df <- readRDS(path) |>
    st_make_valid() |>
    as.data.frame() |>
    filter(year > year_min, year < year_max)

  ids_all_years <- df |>
    distinct(LABEL_ID, year) |>
    count(LABEL_ID, name = "nyears") |>
    filter(nyears == n_distinct(df$year)) |>
    pull(LABEL_ID)

  GBR_ID_reefs <- df |>
    filter(LABEL_ID %in% ids_all_years) |>
    select(LABEL_ID, year, max_aerial_score)

  formatted_data <- GBR_ID_reefs |>
    rename(subject = LABEL_ID, response = max_aerial_score, survey = year) |>
    group_by(survey, subject, response) |>
    summarise(freq = n(), .groups = "drop") |>
    mutate(
      survey   = factor(survey, levels = sort(unique(df$year))),
      response = factor(as.character(response), levels = as.character(0:5))
    )

  ggplot(
    formatted_data,
    aes(
      x = survey,
      stratum = response,
      alluvium = subject,
      y = freq,
      fill = response,
      label = response
    )
  ) +
    geom_flow(color = "black", fill = "lightgrey") +
    geom_stratum(alpha = 0.8) +
    geom_text(stat = "stratum", size = 3) +
    scale_fill_manual(values = interpcolors) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = "", y = expression(italic(n)~"reefs"))
}

map_alluvial <- function(path,
                           year_min = -Inf,
                           year_max = Inf) {
 
  
    df <- readRDS(path) |>
    st_make_valid() |> 
    filter(year > year_min, year < year_max)
  
  ids_all_years <- df |>
    distinct(LABEL_ID, year) |>
    count(LABEL_ID, name = "nyears") |>
    filter(nyears == n_distinct(df$year)) |>
    pull(LABEL_ID) 
  
  GBR_ID_reefs <- df |>
    filter(LABEL_ID %in% ids_all_years) |>
    select(LABEL_ID, year, max_aerial_score)
  
  
  ausmapcrop <- readRDS("~/GBR-dhw/data/ausmap.rds") |> 
    st_transform(crs(df)) |> st_crop(st_bbox(GBR_ID_reefs |> st_buffer(50000)))
  
  formatted_data <- GBR_ID_reefs |>
    rename(subject = LABEL_ID, response = max_aerial_score, survey = year) |>
    group_by(survey, subject, response) |>
    summarise(freq = n(), .groups = "drop") |>
    mutate(
      survey   = factor(survey, levels = sort(unique(df$year))),
      response = factor(as.character(response), levels = as.character(0:5))
    ) |> st_centroid()
  
  interpcolors <- c("#4DADE6", "#5BC36E", "#FFFF80",
                    "#FF784E", "#D53B50", "#5A0E0E")
  
  formatted_data |># filter(year=="1998") |> 
    ggplot() + theme_bw() +
    geom_sf(aes(fill=as.character(as.numeric(response)-1)), shape=21, alpha=0.7, size=3) +
    scale_fill_manual(values=interpcolors, name="aerial\nscore") +
    geom_sf(data = ausmapcrop, fill = "#E2E7D4", alpha = 1) +
    facet_wrap(~survey, nrow=1) +
    coord_sf(lims_method = "geometry_bbox") 

}


figSa <- plot_alluvial(
  path = "~/GBR-dhw/datasets/gbr_shape_aerial_OISST.rds",
  year_min = -Inf,
  year_max = 2026
)

figSb <- map_alluvial(
  path = "~/GBR-dhw/datasets/gbr_shape_aerial_OISST.rds",
  year_min = -Inf,
  year_max = 2026
)


figSa1 <- plot_alluvial(
  path = "~/GBR-dhw/datasets/gbr_shape_aerial_OISST.rds",
  year_min = 2016,
  year_max = 2026
)

figSb1 <- map_alluvial(
  path = "~/GBR-dhw/datasets/gbr_shape_aerial_OISST.rds",
  year_min = 2016,
  year_max = 2026
)

wrap_elements((figSa/figSb) + plot_annotation(title = "1998-2025 (n=30)")) /
  wrap_elements((figSa1/figSb1) +  plot_annotation(title = "2016-2025 (n=81)")) +
  plot_layout(widths = c(1, 1, 1, 1)) +
  plot_annotation(
    tag_levels = "A",
    tag_prefix = "(",
    tag_suffix = ")"
  ) 

```

### Figure S13 Within-reef change in bleaching severity 

For reefs surveyed in all events, we regressed the change in aerial bleaching score on the change in DHW between each pair of events; the intercept (Δscore at ΔDHW = 0) estimates the change in bleaching a reef would show at identical thermal stress. Between consecutive events this matched-heat intercept was near zero or negative (e.g. 2016→2017, −0.35 [−0.63, −0.06]; 2022→2024, −0.63 [−0.94, −0.32]), consistent with recovery dominating short intervals. Across long baselines spanning the pre-/post-2016 thermal regime boundary it was consistently positive: a reef experiencing the same DHW in 2020–2022 as in 1998–2002 bleached approximately one to one-and-a-half ordered categories more severely two decades later (2002→2022, +1.68 [1.38, 1.98]; 2002→2020, +1.13 [0.70, 1.56]). Cells in which the within-reef ΔDHW distribution did not span zero are marked as extrapolated and not interpreted. Because each contrast is identified from repeated observation of the same reefs, this increase cannot be attributed to between-year differences in survey composition.

```{r}
#| message: false
#| warning: false
#| cache: true
#| fig-height: 12
#| fig-width: 12
#| 
year_order <- c("1998", "2002", "2016","2017","2020","2022","2024")
pair_levels <- paste(head(year_order,-1), tail(year_order,-1), sep = "→")

fd5 <- GBRMP_aerial_centroids |>
  st_drop_geometry() |>
  filter(year %in% year_order) |>
  mutate(year = factor(year, levels = year_order)) |>
  # guard against duplicate reef-year rows that would break diff():
  group_by(LABEL_ID, year) |>
  summarise(dhw = mean(dhw), aerial = mean(aerial), .groups = "drop") |>
  group_by(LABEL_ID) |>
  filter(n_distinct(year) == 7) |>            # BALANCED: reefs present in all 5
  arrange(year, .by_group = TRUE) |>
  reframe(                                    # one row per consecutive pair
    year_pair = pair_levels,
    ddhw      = diff(dhw),
    dscore    = diff(aerial)
  ) |>
  mutate(year_pair = factor(year_pair, levels = pair_levels)) |> 
  mutate(year = str_extract(as.character(year_pair), "^\\d{4}") |> as.integer())

cell_stats5 <- fd5 |>
  group_by(year_pair) |>
  filter(n() >= 5) |>
  group_modify(~{
    m  <- lm(dscore ~ ddhw, data = .x)
    ci <- confint(m)["(Intercept)", ]
    tibble(
      b0  = coef(m)[["(Intercept)"]],
      lo  = ci[1], hi = ci[2],
      n   = nrow(.x),
      straddles0 = min(.x$ddhw) < 0 & max(.x$ddhw) > 0,
      gap_sd     = abs(mean(.x$ddhw)) / sd(.x$ddhw)
    )
  }) |>
  ungroup() |>
  mutate(
    extrap = !straddles0 | gap_sd > 1.5,
    sig    = !extrap & (lo > 0 | hi < 0),
    label  = sprintf("%.2f [%.2f, %.2f]%s", b0, lo, hi, ifelse(extrap, "*", ""))
  )

length(unique(fd5$LABEL_ID))   # <-- balanced-panel reef count; expect < 80

repeat_a <- ggplot(fd5, aes(ddhw, dscore)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_point(aes(fill = dscore), alpha = 0.7, shape=21, size = 2,
             show.legend = FALSE) +
  geom_smooth(method = "lm", se = TRUE, colour = "black") +            # free intercept
  geom_text(data = cell_stats5,
            aes(x = -Inf, y = Inf, label = label, colour = sig),
            hjust = -0.06, vjust = 1.4, size = 4.5, fontface = "bold",
            inherit.aes = FALSE) +
  
  scale_colour_manual(values = c(`TRUE` = "#1a6e3c", `FALSE` = "darkred"),
                      guide = "none") +
  # geom_smooth(method = "lm", formula = y ~ 0 + x, se = FALSE,
  #             colour = "steelblue", linetype = "dashed") +            # origin-forced
  facet_wrap(~ year_pair, nrow = 1) +
  scale_x_continuous(limits = c(-10,10), breaks = seq(-10,10,4)) +
  scale_y_continuous(limits = c(-4,4), breaks = seq(-4,4,1)) +
  labs(x = "Δ DHW (later − earlier event)",
       y = "Δ aerial score (later − earlier)") +
  theme_bw() +
  scale_fill_distiller(palette="RdBu", name = "Δ aerial") 


# ggplot() + theme_bw() +
#   geom_point(data = fd5, aes(year, dscore),
#             alpha = 0.7, size = 2,
#              show.legend = FALSE) +
#   geom_smooth(data = fd5, aes(year, dscore),
#               method = "lm", se = TRUE, colour = "black") +            # free intercept
#   # geom_smooth(method = "lm", formula = y ~ 0 + x, se = FALSE,
#   #             colour = "steelblue", linetype = "dashed") +            # origin-forced
#   facet_wrap(~ year_pair, nrow = 1) +
#   scale_x_continuous(limits = c(-10,10), breaks = seq(-10,10,4)) +
#   scale_y_continuous(limits = c(-4,4), breaks = seq(-4,4,1)) +
#   labs(x = "Δ DHW (later − earlier event)",
#        y = "Δ aerial score (later − earlier)") +
#   theme_bw()
# 
# 
# early <- c("1998","2002"); late <- c("2020","2022","2024")
# 
# long_fd <- GBRMP_aerial_centroids |>
#   st_drop_geometry() |>
#   filter(year %in% c(early, late)) |>
#   group_by(LABEL_ID, year) |>
#   summarise(dhw = mean(dhw), aerial = mean(aerial), .groups = "drop") |>
#   group_by(LABEL_ID) |>
#   filter(any(year %in% early) & any(year %in% late)) |>
#   summarise(
#     # earliest available early event vs latest available late event
#     y0 = year[year %in% early][which.min(as.integer(as.character(year[year %in% early])))],
#     y1 = year[year %in% late ][which.max(as.integer(as.character(year[year %in% late ])))],
#     ddhw   = dhw[year == y1]    - dhw[year == y0],
#     dscore = aerial[year == y1] - aerial[year == y0],
#     .groups = "drop") |>
#   mutate(year_pair = paste(y0, y1, sep = "→"))
# 
# nrow(long_fd)   # long-baseline reef count — the number that decides this
# 
# 
# ggplot(long_fd, aes(ddhw, dscore)) +
#   geom_hline(yintercept = 0, colour = "grey60") +
#   geom_vline(xintercept = 0, colour = "grey60") +
#   geom_point(aes(colour = year_pair), alpha = 0.7, size = 2,
#              show.legend = FALSE) +
#   geom_smooth(method = "lm", se = TRUE, colour = "black") +            # free intercept
#   # geom_smooth(method = "lm", formula = y ~ 0 + x, se = FALSE,
#   #             colour = "steelblue", linetype = "dashed") +            # origin-forced
#   facet_wrap(~ year_pair, nrow = 2) +
#   scale_x_continuous(limits = c(-8,8), breaks = seq(-8,8,2)) +
#   scale_y_continuous(limits = c(-4,4), breaks = seq(-4,4,1)) +
#   labs(x = "Δ DHW (later − earlier event)",
#        y = "Δ aerial score (later − earlier)") +
#   theme_bw()




year_order <- c(1998, 2002, 2016, 2017, 2020, 2022, 2024)

# balanced long table: reefs present in all 7 events, one row per reef-year
bal <- GBRMP_aerial_centroids |>
  st_drop_geometry() |>
  filter(year %in% year_order) |>
  group_by(LABEL_ID, year) |>
  summarise(dhw = mean(dhw), aerial = mean(aerial), .groups = "drop") |>
  mutate(year = as.integer(as.character(year))) |>
  group_by(LABEL_ID) |>
  filter(n_distinct(year) == 7) |>
  ungroup()

# all ordered pairs (early < late) via self-join on reef
fd_grid <- bal |>
  rename(y0 = year, dhw0 = dhw, aerial0 = aerial) |>
  inner_join(bal |> rename(y1 = year, dhw1 = dhw, aerial1 = aerial),
             by = "LABEL_ID", relationship = "many-to-many") |>
  filter(y0 < y1) |>
  transmute(LABEL_ID,
            y0 = factor(y0, levels = year_order),
            y1 = factor(y1, levels = year_order),
            ddhw   = dhw1   - dhw0,
            dscore = aerial1 - aerial0)

length(unique(fd_grid$LABEL_ID))   # balanced-7 reef count

library(broom)

# ---- per-cell intercept at ΔDHW = 0, with support diagnostics --------------
cell_stats <- fd_grid |>
  group_by(y0, y1) |>
  filter(n() >= 5) |>                                   # need enough points to fit
  group_modify(~{
    m  <- lm(dscore ~ ddhw, data = .x)
    ci <- confint(m)["(Intercept)", ]
    tibble(
      b0      = coef(m)[["(Intercept)"]],               # matched-heat Δscore
      lo      = ci[1],
      hi      = ci[2],
      n       = nrow(.x),
      # does the ΔDHW cloud actually straddle 0? if not, b0 is extrapolation
      straddles0 = min(.x$ddhw) < 0 & max(.x$ddhw) > 0,
      # how far is 0 from the data centre, in SDs of ddhw (leverage proxy)
      gap_sd  = abs(0 - mean(.x$ddhw)) / sd(.x$ddhw)
    )
  }) |>
  ungroup() |>
  mutate(
    extrap = !straddles0 | gap_sd > 1.5,                # flag shaky intercepts
    label  = sprintf("%.2f [%.2f, %.2f]%s",
                     b0, lo, hi, ifelse(extrap, "*", "")),
    # colour the label by whether the CI clears zero (and isn't extrapolated)
    sig    = !extrap & (lo > 0 | hi < 0)
  )
library(tidyverse)
library(grid)

panel_bg <- expand_grid(
  y0 = levels(fd_grid$y0),
  y1 = levels(fd_grid$y1)
) |>
  mutate(
    y0_num = as.integer(as.character(y0)),
    y1_num = as.integer(as.character(y1)),
    non_year = y1_num <= y0_num
  ) |>
  filter(non_year)

repeat_b <- ggplot() +
  geom_rect(
    data = panel_bg,
    aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
    fill = "grey90",
    colour = NA,
    inherit.aes = FALSE
  ) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_point(
    data = fd_grid,
    aes(ddhw, dscore, fill = dscore),
    alpha = 0.5,
    shape = 21,
    size = 1.3,
    stroke = 0.1
  ) +
  geom_smooth(
    data = fd_grid,
    aes(ddhw, dscore),
    method = "lm",
    se = TRUE,
    colour = "black",
    linewidth = 0.6
  ) +
  geom_text(
    data = cell_stats,
    aes(
      x = -Inf,
      y = Inf,
      label = label,
      colour = sig
    ),
    hjust = -0.08,
    vjust = 1.4,
    size = 2.7,
    fontface = "bold",
    inherit.aes = FALSE
  ) +
  scale_colour_manual(
    values = c(`TRUE` = "#1a6e3c", `FALSE` = "darkred"),
    guide = "none"
  ) +
  geom_text(
    data = cell_stats,
    aes(
      x = Inf,
      y = -Inf,
      label = paste0("n=", n)
    ),
    hjust = 1.1,
    vjust = -0.6,
    size = 2.5,
    colour = "grey45",
    inherit.aes = FALSE
  ) +
  facet_grid(y0 ~ y1, drop = FALSE) +
  scale_x_continuous(limits = c(-10, 10), breaks = seq(-8, 8, 4)) +
  scale_y_continuous(limits = c(-4, 5), breaks = seq(-4, 4, 2)) +
  scale_fill_distiller(palette = "RdBu", name = "Δ aerial") +
  labs(
    x = "Δ DHW (later − earlier event)",
    y = "Δ aerial score (later − earlier)"
  ) +
  theme_bw() +
  theme(
    panel.spacing = unit(2, "pt")
  )

repeat_a / repeat_b + plot_layout(heights = c(0.2, 1))


```

### Figure S14 Latitudinal trends in bleaching 

Latitudinal distribution of aerial bleaching severity across the Great Barrier Reef, pooled across survey years. Bars represent the total number of reefs within each 1° latitude bin, with colours indicating the maximum aerial bleaching score recorded for each reef. Figure summarises the cumulative spatial pattern of bleaching observations rather than interannual variation. Reef counts are highest through the central GBR, particularly around 21–22°S and 20–21°S, where all bleaching categories are represented. Northern and southern latitude bins contain fewer reefs, and therefore smaller total counts, but still show a mixture of bleaching severities.

```{r}
#| message: false
#| warning: false
#| cache: false
#| fig-height: 9
#| fig-width: 8

library(dplyr)
library(sf)
library(ggplot2)

all_events_year_centroid <- gbr_reefs %>%
  st_make_valid() %>%
  mutate(
    area = st_area(geometry)
  ) %>%
  st_centroid() %>%
  mutate(
    lat = st_coordinates(geometry)[, "Y"]
  ) %>%
  st_drop_geometry() %>%
  select(LABEL_ID, lat, area)

all_events_year_lat <- readRDS("/Users/rof011/GBR-dhw/datasets/gbr_shape_aerial_OISST.rds") %>%
  mutate(
    aerial_score = as.character(max_aerial_score)
  ) %>%
  select(-median_aerial_score, -max_aerial_score) %>%
  as.data.frame() %>%
  left_join(all_events_year_centroid, by = "LABEL_ID") %>%
  mutate(
    latbin = floor(lat),
    latbin_lab = paste0(abs(latbin + 1), "–", abs(latbin), "°S"),
    aerial_score = factor(
      aerial_score,
      levels = c("0", "1", "2", "3", "4", "5")
    )
  ) %>%
  group_by(latbin, latbin_lab, aerial_score) %>%
  summarise(
    count = n(),
    area_km2 = as.numeric(sum(area, na.rm = TRUE)) / 1e6,
    .groups = "drop"
  ) %>%
  mutate(
    latbin_lab = factor(
      latbin_lab,
      levels = unique(latbin_lab[order(latbin, decreasing = TRUE)])
    )
  ) %>%
  arrange(desc(latbin), aerial_score)

ggplot(
  all_events_year_lat,
  aes(
    x = count,
    y = latbin_lab,
    fill = aerial_score
  )
) +
  theme_bw() +
  geom_col(
    position = "stack",
    colour = "black",
    linewidth = 0.2
  ) +
  scale_fill_manual(
    name = "",
    values = c(
      "0" = "#ABD9E9",
      "1" = "#FFFFBF",
      "2" = "#FEE090",
      "3" = "#FDAE61",
      "4" = "#F46D43",
      "5" = "#D73027"
    ),
    labels = c(
      "No bleaching (<1%)",
      "Low bleaching (1–10%)",
      "Medium bleaching (11–30%)",
      "High bleaching (31–60%)",
      "Very high bleaching (61–90%)",
      "Extreme bleaching (>91%)"
    )
  ) +
  labs(
    x = expression(italic(n) ~ "reefs"),
    y = "Latitude bin"
  ) +
  guides(
    fill = guide_legend(reverse = TRUE)
  ) +
  theme(
    legend.position = c(0.78, 0.22),
    legend.background = element_rect(fill = "white", colour = "grey70"),
    panel.grid.minor = element_blank()
  )

```

### Figure S15 A decade of severe bleaching on the Great Barrier Reef (2016-2018)

Reef-level maximum bleaching score. Surveyed reefs across the GBR, coloured by maximum aerial bleaching score (0 = none to 5 = most severe) over a satellite basemap, shown in three latitudinal strips (north, central, south). Severe scores concentrate along the northern and central reef matrix, with a localised severe cluster in the far south, mirroring the latitudinal thermal-stress structure in Fig. S3.

```{r}
#| message: false
#| warning: false
#| cache: false
#| fig-height: 15.5
#| fig-width: 4.5


maxOISST <- readRDS("/Users/rof011/GBR-dhw/datasets/gbr_shape_aerial_OISST.rds") |>
  group_by(LABEL_ID) |>
  filter(year > 2016) |> 
  slice_max(max_aerial_score, n=1) |>
  drop_na(max_aerial_score)


maxOISST_faceted <- maxOISST |>
  st_make_valid() |>
  mutate(
    region = case_when(
      st_coordinates(st_centroid(geometry))[, "Y"] > -15.5 ~ "North",
      st_coordinates(st_centroid(geometry))[, "Y"] > -20.5 ~ "Central",
      TRUE ~ "South"
    ),
    region = factor(region, levels = c("North", "Central", "South"))
  )

# north_plot <- ggplot() +
#   theme_bw() +
#   annotation_map_tile(
#     type = "esri",
#     zoomin = 0
#   ) +
#   geom_sf(
#     data = maxOISST_faceted |> filter(region=="North"),
#     aes(fill = factor(max_aerial_score)),
#     shape = 21,
#     colour = "grey20",
#     linewidth = 0.1,
#     show.legend=TRUE
#   ) +
#   scale_fill_brewer(
#     palette = "RdBu",
#     direction = -1,
#     na.value = "grey80",
#     name = "Max aerial score"
#   ) +
#   labs(
#     x = NULL,
#     y = NULL
#   ) +
#   theme(
#     legend.position = "inside",
#     legend.position.inside = c(0.78, 0.78),
#     legend.justification = c(1, 0),
#     legend.background = element_rect(
#       fill = scales::alpha("white", 0.8),
#       colour = "grey40",
#       linewidth = 0.3
#     ),
#     legend.key = element_rect(fill = NA, colour = NA),
#     strip.background = element_rect(fill = "grey95", colour = "grey40"),
#     strip.text = element_text(face = "bold")
#   ) +
#   scale_colour_manual(values = c("#ABD9E9", "#FFFFBF", "#FEE090", "#FDAE61", "#F46D43",  "#D73027"), name = "") +
#   scale_fill_manual(values = c("#ABD9E9", "#FFFFBF", "#FEE090", "#FDAE61", "#F46D43",  "#D73027"), name = "") +
#   geom_sf(data = ausmap |> st_crop(maxOISST_faceted |> filter(region=="North") |> st_bbox()), fill = "#E2E7D4", alpha = 1)
#
# central_plot <- ggplot() +
#   theme_bw() +
#   annotation_map_tile(
#     type = "esri",
#     zoomin = 0
#   ) +
#   geom_sf(
#     data = maxOISST_faceted |> filter(region=="Central"),
#     aes(fill = factor(max_aerial_score)),
#     shape = 21,
#     colour = "grey20",
#     linewidth = 0.1,
#     show.legend=FALSE
#   ) +
#   scale_fill_brewer(
#     palette = "RdBu",
#     direction = -1,
#     na.value = "grey80",
#     name = "Max aerial score"
#   ) +
#   labs(
#     x = NULL,
#     y = NULL
#   ) +
#   theme(
#     legend.position = "right",
#     strip.background = element_rect(fill = "grey95", colour = "grey40"),
#     strip.text = element_text(face = "bold")
#   ) +
#   scale_colour_manual(values = c("#ABD9E9", "#FFFFBF", "#FEE090", "#FDAE61", "#F46D43",  "#D73027"), name = "") +
#   scale_fill_manual(values = c("#ABD9E9", "#FFFFBF", "#FEE090", "#FDAE61", "#F46D43",  "#D73027"), name = "") +
#   geom_sf(data = ausmap |> st_crop(maxOISST_faceted |> filter(region=="Central") |> st_bbox()), fill = "#E2E7D4", alpha = 1)
#
#
# south_plot <- ggplot() +
#   theme_bw() +
#   annotation_map_tile(
#     type = "esri",
#     zoomin = 0
#   ) +
#   geom_sf(
#     data = maxOISST_faceted |> filter(region=="South"),
#     aes(fill = factor(max_aerial_score)),
#     shape = 21,
#     colour = "grey20",
#     linewidth = 0.1,
#     show.legend=FALSE
#   ) +
#   scale_fill_brewer(
#     palette = "RdBu",
#     direction = -1,
#     na.value = "grey80",
#     name = "Max aerial score"
#   ) +
#   labs(
#     x = NULL,
#     y = NULL
#   ) +
#   theme(
#     legend.position = "right",
#     strip.background = element_rect(fill = "grey95", colour = "grey40"),
#     strip.text = element_text(face = "bold")
#   ) +
#   scale_colour_manual(values = c("#ABD9E9", "#FFFFBF", "#FEE090", "#FDAE61", "#F46D43",  "#D73027"), name = "") +
#   scale_fill_manual(values = c("#ABD9E9", "#FFFFBF", "#FEE090", "#FDAE61", "#F46D43",  "#D73027"), name = "") +
#   geom_sf(data = ausmap |> st_crop(maxOISST_faceted |> filter(region=="South") |> st_bbox()), fill = "#E2E7D4", alpha = 1)
#
#
#
# (north_plot + labs(title = NULL)) |
#   (central_plot + labs(title = NULL)) /
#   (south_plot + labs(title = NULL)) #+
#   # plot_layout(guides = "collect") &
#   # theme(
#   #   legend.position = "bottom"
#   # )

library(tidyverse)
library(sf)
library(tmap)

tmap_mode("plot")

score_cols <- c(
  "0" = "#ABD9E9",
  "1" = "#FFFFBF",
  "2" = "#FEE090",
  "3" = "#FDAE61",
  "4" = "#F46D43",
  "5" = "#D73027"
)

maxOISST_faceted <- maxOISST_faceted |>
  mutate(
    region = factor(region, levels = c("North", "Central", "South")),
    max_aerial_score = factor(max_aerial_score, levels = 0:5)
  )

ausmapoutline <- rnaturalearth::ne_countries(scale="large", type = "countries") |>
  st_transform(4326)


north_map <- tm_basemap("Esri.WorldImagery", alpha=0.6) +
  tm_shape(ausmapoutline) +
    tm_polygons(
      fill = "#E2E7D4",
      fill_alpha = 0.5,
      col = NA
    ) +
  tm_shape(maxOISST_faceted |> filter(region=="North"), is.main = TRUE) +
    tm_polygons(
      fill = "max_aerial_score",
      fill.scale = tm_scale_categorical(
        values = score_cols,
        value.na = "grey80"),
      lwd = 0.4,
      col = "black"
      ) +
  tm_layout(
    frame = TRUE,
    legend.outside = FALSE,
    legend.position = c("right", "top"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.8,
    inner.margins = c(0, 0.2, 0, 0.2),
    outer.margins = c(0.1, 0.2, 0.1, 0.2)
  )


central_map <- tm_basemap("Esri.WorldImagery", alpha=0.6) +
  tm_shape(ausmapoutline) +
  tm_polygons(
    fill = "#E2E7D4",
    fill_alpha = 0.5,
    col = NA
  ) +
  tm_shape(maxOISST_faceted |> filter(region=="Central"), is.main = TRUE) +
  tm_polygons(
    fill = "max_aerial_score",
    fill.legend = tm_legend_hide(),
    fill.scale = tm_scale_categorical(
      values = score_cols,
      value.na = "grey80"),
    lwd = 0.4,
    col = "black"
  ) +
  tm_layout(
    frame = TRUE,
    legend.outside = FALSE,
    legend.position = c("right", "top"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.8,
    inner.margins = c(0, 0.2, 0, 0.2),
    outer.margins = c(0.1, 0.2, 0.1, 0.2)
  )


south_map <- tm_basemap("Esri.WorldImagery", alpha=0.6) +
  tm_shape(ausmapoutline) +
  tm_polygons(
    fill = "#E2E7D4",
    fill_alpha = 0.5,
    col = NA
  ) +
  tm_shape(maxOISST_faceted |> filter(region=="South"), is.main = TRUE) +
  tm_polygons(
    fill = "max_aerial_score",
    fill.legend = tm_legend_hide(),
    fill.scale = tm_scale_categorical(
      values = score_cols,
      value.na = "grey80"),
    lwd = 0.4,
    col = "black"
  ) +
  tm_layout(
    frame = TRUE,
    legend.outside = FALSE,
    legend.position = c("right", "top"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.8,
    inner.margins = c(0, 0.2, 0, 0.2),
    outer.margins = c(0.1, 0.2, 0.1, 0.2)
  )

#c(bottom, left, top, right)


output2 <- tmap_arrange(
  north_map,
  central_map,
  south_map,
  ncol = 1
)

output2

```

### Figure S16 Latitudinal trends in bleaching across events

Temporal and latitudinal structure of aerial bleaching severity across individual GBR reefs. Each row represents a reef ordered from north to south by centroid latitude, and each column represents a survey year. Tile colour indicates the maximum aerial bleaching score recorded for that reef in that year, ranging from no bleaching to extreme bleaching. 

```{r}
#| message: false
#| warning: false
#| cache: false
#| fig-height: 15.5
#| fig-width: 4.5

library(dplyr)
library(tidyr)
library(sf)
library(ggplot2)

event_years_required <- c(1998, 2002, 2016, 2017, 2020, 2022, 2024)
event_years_plot <- c(1998, 2002, 2016, 2017, 2022, 2020, 2024, 2025)

latreefs <- readRDS("/Users/rof011/GBR-dhw/datasets/gbr_shape_aerial_OISST.rds") |> st_make_valid()

reef_order <- latreefs %>%
  filter(year %in% event_years_required) %>%
  st_drop_geometry() %>%
  group_by(LABEL_ID) %>%
  filter(all(event_years_required %in% year)) %>%
  ungroup() %>%
  distinct(LABEL_ID) %>%
  left_join(
    latreefs %>%
      st_centroid() %>%
      mutate(lat = st_coordinates(geometry)[, "Y"]) %>%
      st_drop_geometry() %>%
      group_by(LABEL_ID) %>%
      summarise(lat = mean(lat, na.rm = TRUE), .groups = "drop"),
    by = "LABEL_ID"
  ) %>%
  arrange((lat)) %>%
  pull(LABEL_ID)

latreefs_plot <- latreefs %>%
  filter(
    year %in% event_years_plot,
    LABEL_ID %in% reef_order
  ) %>%
  st_drop_geometry() %>%
  select(LABEL_ID, year, max_aerial_score) %>%
  mutate(
    max_aerial_score = as.character(max_aerial_score)
  ) %>%
  complete(
    LABEL_ID = reef_order,
    year = event_years_plot
  ) %>%
  mutate(
    LABEL_ID = factor(LABEL_ID, levels = reef_order),
    year = factor(year, levels = event_years_plot),
    max_aerial_score = factor(
      max_aerial_score,
      levels = c("0", "1", "2", "3", "4", "5")
    )
  )

ggplot(
  latreefs_plot,
  aes(
    x = year,
    y = LABEL_ID,
    fill = max_aerial_score
  )
) +
  theme_bw() +
  geom_tile(color="black", linewidth=0.05) +
  scale_fill_manual(
    name = "",
    values = c(
      "0" = "#ABD9E9",
      "1" = "#FFFFBF",
      "2" = "#FEE090",
      "3" = "#FDAE61",
      "4" = "#F46D43",
      "5" = "#D73027"
    ),
    labels = c(
      "No bleaching (<1%)",
      "Low bleaching (1–10%)",
      "Medium bleaching (11–30%)",
      "High bleaching (31–60%)",
      "Very high bleaching (61–90%)",
      "Extreme bleaching (>91%)"
    ),
    na.value = "grey90",
    drop = FALSE
  ) +
  labs(
    x = "Year",
    y = ""
  ) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid = element_blank()
  ) +
  guides(fill = guide_legend(reverse = TRUE))

```



## Supplemenentary Tables

### Table S1


```{r}
#| message: false
#| warning: false
library(tidyverse); library(kableExtra); library(INLA)
 
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
 
fm1 <- readRDS("~/GBR-dhw/data/fm1.rds")


# ---- 1. Fixed effects -------------------------------------------------------
fe_labels <- c(
  "(Intercept)"  = "Intercept (baseline, 1998)",
  "dhw"          = "DHW (per °C-week)",
  "year2002"="Year: 2002", "year2016"="Year: 2016", "year2017"="Year: 2017",
  "year2020"="Year: 2020", "year2022"="Year: 2022", "year2024"="Year: 2024",
  "year2025"="Year: 2025",
  "dhw:year2002"="DHW × 2002","dhw:year2016"="DHW × 2016","dhw:year2017"="DHW × 2017",
  "dhw:year2020"="DHW × 2020","dhw:year2022"="DHW × 2022","dhw:year2024"="DHW × 2024",
  "dhw:year2025"="DHW × 2025")
 
fixed_tab <- fm1$summary.fixed |>
  rownames_to_column("term") |>
  transmute(Parameter = fe_labels[term],
            Mean = fmt(mean),
            SD = fmt(sd),
            `2.5%` = fmt(`0.025quant`),
            `Median` = fmt(`0.5quant`),
            `97.5%` = fmt(`0.975quant`)) |>
  mutate(group = case_when(grepl("^DHW ×", Parameter) ~ "DHW × year interactions",
                           grepl("^Year", Parameter)  ~ "Year effects (vs 1998)",
                           TRUE ~ "Main effects")) |>
  arrange(factor(group, levels=c("Main effects","Year effects (vs 1998)",
                                 "DHW × year interactions")))
 
# ---- 2. Hyperparameters: reconstruct interpretable scales -------------------
hp <- fm1$summary.hyperpar
# ordered POM cut-points from increment thetas
th <- hp[grep("theta.*POM", rownames(hp)), "mean"]
alpha <- cumsum(c(th[1], exp(th[-1])))
# SPDE range/SD: read directly if pcmatern, else transform matern thetas
sp <- tryCatch(inla.spde2.result(fm1, "spatial.field", spde, do.transform = TRUE),
               error = function(e) NULL)
if (!is.null(sp)) {
  rng <- inla.qmarginal(c(.025,.5,.975), sp$marginals.range.nominal[[1]])
  sdv <- sqrt(inla.qmarginal(c(.025,.5,.975), sp$marginals.variance.nominal[[1]]))
} else {  # pcmatern reports natively
  rng <- as.numeric(hp[grep("Range", rownames(hp)), c("0.025quant","0.5quant","0.975quant")])
  sdv <- as.numeric(hp[grep("Stdev", rownames(hp)), c("0.025quant","0.5quant","0.975quant")])
}
iid_prec <- hp[grep("LABEL_ID", rownames(hp)), c("0.025quant","0.5quant","0.975quant")]
iid_sd   <- 1/sqrt(rev(as.numeric(iid_prec)))   # precision -> SD (reverse quantiles)
 
hyper_tab <- tribble(
  ~Parameter, ~`2.5%`, ~Median, ~`97.5%`,
  "POM cut-point α₁ (logit)",            fmt(alpha[1]), "", "",
  "POM cut-point α₂ (logit)",            fmt(alpha[2]), "", "",
  "POM cut-point α₃ (logit)",            fmt(alpha[3]), "", "",
  "Spatial field: range (km)",  fmt(rng[1]/1000,0), fmt(rng[2]/1000,0), fmt(rng[3]/1000,0),
  "Spatial field: marginal SD (logit)", fmt(sdv[1]),   fmt(sdv[2]),     fmt(sdv[3]),
  "Reef random effect: SD (logit)",      fmt(iid_sd[1]),fmt(iid_sd[2]),  fmt(iid_sd[3]))
 
# ---- 3. Model fit -----------------------------------------------------------
fit_tab <- tribble(
  ~Statistic, ~Value,
  "DIC",  fmt(fm1$dic$dic, 1),
  "WAIC", fmt(fm1$waic$waic, 1),
  "Effective parameters (pD)", fmt(fm1$dic$p.eff, 1),
  "Marginal log-likelihood", fmt(fm1$mlik[1], 1))
 
# ---- 4. Render --------------------------------------------------------------
fmt_out <- "html"   # "latex" for manuscript
 
tab_fixed <- fixed_tab |> select(-group) |>
  kbl(format = fmt_out, booktabs = TRUE, align = "lrrrrr",
      caption = "Supplementary Table S1. Posterior summaries of fixed effects from the spatial proportional-odds bleaching model (fm1). Estimates are on the logit scale; year effects are relative to the 1998 reference event.") |>
  kable_styling(latex_options = c("hold_position"), full_width = FALSE,
                bootstrap_options = c("striped","condensed")) |>
  pack_rows(index = table(fixed_tab$group)[unique(fixed_tab$group)])
 
tab_hyper <- hyper_tab |>
  kbl(format = fmt_out, booktabs = TRUE, align = "lrrr",
      caption = "Supplementary Table S1 (cont.). Hyperparameters on interpretable scales: ordered POM cut-points, SPDE spatial range and marginal standard deviation, and reef-level random-effect standard deviation.") |>
  kable_styling(latex_options = c("hold_position"), full_width = FALSE,
                bootstrap_options = c("striped","condensed")) |>
  footnote(general = "Cut-points reconstructed from INLA increment parameterisation (α₁=θ₁; α_k=α_{k-1}+exp(θ_k)). Range in km (domain CRS EPSG:32755). Reef SD derived from the IID precision posterior.",
           general_title = "Note: ", footnote_as_chunk = TRUE)
 
tab_fit <- fit_tab |>
  kbl(format = fmt_out, booktabs = TRUE, align = "lr",
      caption = "Supplementary Table S1 (cont.). Model fit statistics.") |>
  kable_styling(latex_options = c("hold_position"), full_width = FALSE,
                bootstrap_options = c("striped","condensed"))
 
```

```{r}
tab_fixed; 
```

```{r}
tab_hyper
```

```{r}
tab_fit

```

### Table S2

```{r}

library(tidyverse)
library(knitr)
library(kableExtra)
library(broom)

fmt_out <- "html"   # "latex" for manuscript

coef_tab <- tidy(m) |>
  mutate(
    section = "Coefficients",
    term = recode(
      term,
      `(Intercept)` = "Intercept",
      ndjfm_anomaly = "NDJFM anomaly"
    )
  ) |>
  transmute(
    section,
    term,
    estimate = estimate,
    std_error = std.error,
    statistic = statistic,
    p_value = p.value
  )

fit_tab <- glance(m) |>
  tidyr::crossing(
    tibble(
      term = c(
        "Residual standard error",
        "Residual degrees of freedom",
        "R-squared",
        "Adjusted R-squared",
        "F-statistic",
        "Model p-value"
      )
    )
  ) |>
  mutate(
    section = "Model fit",
    estimate = case_when(
      term == "Residual standard error" ~ sigma,
      term == "Residual degrees of freedom" ~ df.residual,
      term == "R-squared" ~ r.squared,
      term == "Adjusted R-squared" ~ adj.r.squared,
      term == "F-statistic" ~ statistic,
      term == "Model p-value" ~ p.value
    ),
    std_error = NA_real_,
    statistic = NA_real_,
    p_value = NA_real_
  ) |>
  select(section, term, estimate, std_error, statistic, p_value)

model_main_tab <- bind_rows(coef_tab, fit_tab) |>
  mutate(
    across(
      c(estimate, std_error, statistic, p_value),
      ~ case_when(
        is.na(.x) ~ NA_character_,
        abs(.x) < 0.001 ~ formatC(.x, format = "e", digits = 2),
        TRUE ~ formatC(.x, format = "f", digits = 3)
      )
    )
  )

tab_model_main <- model_main_tab |>
  kbl(
    format = fmt_out,
    booktabs = TRUE,
    align = "llrrrr",
    col.names = c(
      "Section",
      "Term",
      "Estimate",
      "Std. error",
      "Statistic",
      "P-value"
    ),
    caption = "Supplementary Table S2. Weighted linear model relating year-level bleaching intensity estimates to NDJFM anomaly. Year estimates were weighted by inverse posterior variance."
  ) |>
  kable_styling(
    latex_options = c("hold_position"),
    full_width = FALSE,
    bootstrap_options = c("striped", "condensed")
  ) |>
  pack_rows(index = table(model_main_tab$section)[unique(model_main_tab$section)])

tab_model_main

``
